### 2020-05-04 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
Initial commit

---
title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
   In 2020 Google released a set of Community Mobility Reports (GCMR). These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline. This novel source of data offers an opportunity to investigate the way mobility potentially correlates with transmission of COVID-19. Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US. The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.

journal: "Transportation Findings"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r load-packages, include=FALSE}
#Load the packages needed to read and work with the data
library(adespatial)
library(countrycode)
library(gridExtra)
library(interactions)
library(jtools)
library(kableExtra)
library(lubridate)
library(metR)
library(plotly)
library(RCurl)
library(sf)
library(spdep)
library(tidyverse)
library(tigris)
library(usmap)
library(wppExplorer)
```

```{r gcmr-data-read-and-save, eval=FALSE, include=FALSE}
# Run only to read more recent data
#Retrieve data:
download <- getURL("https://www.gstatic.com/covid19/mobility/Global_Mobility_Report.csv?cachebust=911a386b6c9c230f")
gmobility <- read.csv (text = download)

#Save as RData to keep a local copy of the mobility report:
save(gmobility, file = "gmobility.RData")
```

```{r load-saved-gcmr-data, include=FALSE}
#Load data:
load("gmobility.RData")
```

```{r extract-gcmr-for-us, include=FALSE}
gmobility <- gmobility %>%
  filter(country_region == "United States") %>%
  droplevels()
```

```{r prepare-mobility-report-data, include=FALSE}
#`date` is stored as a factor which is not ideal work with times; Change format:
gmobility <- gmobility %>%
  mutate(date = lubridate::ymd(date)) %>%
  #rename the annoyingly long names of the columns to something more practical
  transmute(date,
            state = as.character(sub_region_1),
         county = as.character(sub_region_2),
         r_and_r = retail_and_recreation_percent_change_from_baseline, 
         g_and_p = grocery_and_pharmacy_percent_change_from_baseline, 
         parks = parks_percent_change_from_baseline,
         transit = transit_stations_percent_change_from_baseline, 
         work = workplaces_percent_change_from_baseline,
         residential = residential_percent_change_from_baseline) %>%
  #convert the mobility indicators to differences from the baseline, so "1" is the baseline value
  mutate(r_and_r = 1 + r_and_r/100,
         g_and_p = 1 + g_and_p/100,
         parks = 1 + parks/100,
         transit = 1 + transit/100,
         work = 1 + work/100,
         residential = 1 + residential/100)
```

```{r filter-state-gcmr-data, include=FALSE}
#To extract the state-level mobility indicators select those records that do not include the label for the county (i.e., sub_region_2 is empty), and that are not national level statistics (i.e., sub_region_1 is not empty)
cmr <- gmobility %>%
  filter(state !="", 
         county == "")
```

```{r nyt-covid-19-data-read-and-save, eval=FALSE, include=FALSE}
# Run only to read more recent data
#retrieve data
c19 <- read.csv(file.path("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"),
                 check.names = FALSE, 
                 stringsAsFactors = TRUE) %>%
  mutate(date = ymd(date))

#Save as RData to keep a local copy of the reports:
save(c19, file = "covid19_nyt.RData")
```

```{r load-saved-covid-19-data, include=FALSE}
#Load data:
load("covid19_nyt.RData")
```

```{r coterminous-states, include=FALSE}
#filter states to keep only the coterminous states
c19 <- c19 %>%
  filter(state != "Hawaii",
         state != "Alaska",
         state != "United States Virgin Islands",
         state != "Commonwealth of the Northern Mariana Islands",
         state != "Guam",
         state != "American Samoa",
         state != "Puerto Rico")
```

```{r add-day-variable, include=FALSE}
#add day variable:
c19 <- c19 %>%
  mutate(day = yday(date) - min(yday(date)))
```

```{r calculate-new-daily-cases, include=FALSE}
#o calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total:
c19 <- c19 %>%
  group_by(state) %>%
  mutate(new_cases = cases - lag(cases, 1)) %>%
  ungroup() %>%
  #there is one record with a negative number of new cases, likely a correction (Georgia in 2020-04-12). Remove record:
  filter(new_cases >= 0)
```

```{r state-population-data, include=FALSE}
#Retrieve population data:
population <- statepop %>%
  transmute(pop_2015, state = full)
```

```{r join-population-to-covid-19-data, include=FALSE}
#Join population to covid-19 data:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(population, by = c("state" = "state"))
```

```{r calculate-incidence, include=FALSE}
#Calculate the incidence of reported cases by 100,000 population:
c19 <- c19 %>%
  mutate(incidence = cases/(pop_2015/100000),
         log_incidence = log(incidence))
```

```{r calculate-log-new-cases, include=FALSE}
#calculate log of new cases
c19 <- c19 %>% 
  mutate(log_new_cases = log(new_cases + 0.001))
```

```{r join-gcmr-and-covid-19-data, include=FALSE}
#Join the tables:
c19 <- c19 %>% 
  mutate(state = as.character(state)) %>%
  left_join(cmr, by = c("state" = "state", "date" = "date"))
```

```{r calculate-lagged-mobility-indicators, include=FALSE}
#The potential salutary effect of the lockdown (measured in terms of reduced mobility) are not immediate, given the incubation period of the disease. The incubation period has been estimated to be between 2 and 11 days (95% interval) by [Lauer et al. (2020)](https://annals.org/aim/fullarticle/2762808/incubation-period-coronavirus-disease-2019-covid-19-from-publicly-reported). If we wanted to explore the way reduced mobility correlates with number of new cases, we would need to lag the mobility variables by some period.

# Lagged retail and recreation: 11 day moving average
r_and_r_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$r_and_r, 12) + 
                      lag(.x$r_and_r, 11) + 
                      lag(.x$r_and_r, 10) + 
                      lag(.x$r_and_r, 9) + 
                      lag(.x$r_and_r, 8) + 
                      lag(.x$r_and_r, 7) + 
                      lag(.x$r_and_r, 6) + 
                      lag(.x$r_and_r, 5) + 
                      lag(.x$r_and_r, 4) + 
                      lag(.x$r_and_r, 3) + 
                      lag(.x$r_and_r, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_r_and_r_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_r_and_r_lag11)

# Lagged groceries and pharmacies: 11 day moving average
g_and_p_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$g_and_p, 12) + 
                      lag(.x$g_and_p, 11) + 
                      lag(.x$g_and_p, 10) + 
                      lag(.x$g_and_p, 9) + 
                      lag(.x$g_and_p, 8) + 
                      lag(.x$g_and_p, 7) + 
                      lag(.x$g_and_p, 6) + 
                      lag(.x$g_and_p, 5) + 
                      lag(.x$g_and_p, 4) + 
                      lag(.x$g_and_p, 3) + 
                      lag(.x$g_and_p, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_g_and_p_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_g_and_p_lag11)

# Lagged parks: 11 day moving average
parks_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$parks, 12) + 
                      lag(.x$parks, 11) + 
                      lag(.x$parks, 10) + 
                      lag(.x$parks, 9) + 
                      lag(.x$parks, 8) + 
                      lag(.x$parks, 7) + 
                      lag(.x$parks, 6) + 
                      lag(.x$parks, 5) + 
                      lag(.x$parks, 4) + 
                      lag(.x$parks, 3) + 
                      lag(.x$parks, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_parks_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_parks_lag11)

# Lagged groceries and pharmacies: 11 day moving average
transit_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$transit, 12) + 
                      lag(.x$transit, 11) + 
                      lag(.x$transit, 10) + 
                      lag(.x$transit, 9) + 
                      lag(.x$transit, 8) + 
                      lag(.x$transit, 7) + 
                      lag(.x$transit, 6) + 
                      lag(.x$transit, 5) + 
                      lag(.x$transit, 4) + 
                      lag(.x$transit, 3) + 
                      lag(.x$transit, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_transit_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_transit_lag11)

# Lagged groceries and pharmacies: 11 day moving average
work_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$work, 12) + 
                      lag(.x$work, 11) + 
                      lag(.x$work, 10) + 
                      lag(.x$work, 9) + 
                      lag(.x$work, 8) + 
                      lag(.x$work, 7) + 
                      lag(.x$work, 6) + 
                      lag(.x$work, 5) + 
                      lag(.x$work, 4) + 
                      lag(.x$work, 3) + 
                      lag(.x$work, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_work_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_work_lag11)

# Lagged residential: 11 day moving average
residential_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$residential, 12) + 
                      lag(.x$residential, 11) + 
                      lag(.x$residential, 10) + 
                      lag(.x$residential, 9) + 
                      lag(.x$residential, 8) + 
                      lag(.x$residential, 7) + 
                      lag(.x$residential, 6) + 
                      lag(.x$residential, 5) + 
                      lag(.x$residential, 4) + 
                      lag(.x$residential, 3) + 
                      lag(.x$residential, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_residential_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_residential_lag11)
```

```{r join-lagged-indicators, include=FALSE}
#join lagged mobility indicators to table:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(r_and_r_lag11, by = c("state" = "state", "date" = "date")) %>%
  left_join(g_and_p_lag11, by = c("state" = "state", "date" = "date"))  %>%
  left_join(parks_lag11, by = c("state" = "state", "date")) %>%
  left_join(transit_lag11, by = c("state" = "state", "date")) %>%
  left_join(work_lag11, by = c("state" = "state", "date")) %>%
  left_join(residential_lag11, by = c("state" = "state", "date"))
```

```{r remove-nas, include=FALSE}
#Remove all the "NAs" to obtain a clean table with COVID-19 and mobility information:
c19 <- c19 %>%
  drop_na()
```

```{r prepare-variables-for-analysis, include=FALSE}
#replace mobility indicators by their lagged moving averages
c19 <- c19 %>% 
  mutate(retail = Mean_r_and_r_lag11,
         groceries = Mean_g_and_p_lag11,
         parks = Mean_parks_lag11,
         parks2 = Mean_parks_lag11^2,
         transit = Mean_transit_lag11,
         work = Mean_work_lag11,
         residential = Mean_residential_lag11,
         residential2 = Mean_residential_lag11^2,
         parksXresidential = parks * residential,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NY = ifelse(state == "New York", 1, 0),
         NYxday = NY * day)
```

Research Questions and Hypotheses
==========================

The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders. Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}. These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.

This paper investigates the potential of Google Community Mobility Reports to asses the impact of mobility on COVID-19. The following questions are posed:

- Do changes in mobility according to GCMR correlate with the transmission of COVID-19?
- And if so, what do we learn about mobility and the spread of the disease?

The source document for this paper is an `R` markdown document available from the following repository:

https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US

Methods and Data
============

GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places, namely retail and recreation, groceries and pharmacies, parks, transit stations, workplaces, and residential. Mobility indicators for each of these places are reported as percentage changes from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.

For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change. The incubation time of the disease has been estimated to be between 2 and 12 days (95% interval) by Lauer et al. [-@Lauer2020incubation]. Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases. For this reason, lagged moving averages of the mobility indicators are calculated. Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence. Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.

```{r descriptive-statistics, echo=FALSE}
data.frame(Variable = c("New Cases",
                        "date",
                        "retail",
                        "groceries",
                        "parks",
                        "transit",
                        "work",
                        "residential"),
           Definition = c("New Daily Cases of COVID-19",
                        "Date",
                        "Retail and Recreation",
                        "Groceries and Pharmacies",
                        "Parks",
                        "Transit Stations",
                        "Workplaces",
                        "Residential"),
           min = c(as.character(min(c19$new_cases)),
                   as.character(min(c19$date)),
                   as.character(round(min(c19$retail), 2)),
                   as.character(round(min(c19$groceries), 2)),
                   as.character(round(min(c19$parks), 2)),
                   as.character(round(min(c19$transit), 2)),
                   as.character(round(min(c19$work), 2)),
                   as.character(round(min(c19$residential), 2))),
           median = c(as.character(median(c19$new_cases)),
                   as.character(median(c19$date)),
                   as.character(round(median(c19$retail), 2)),
                   as.character(round(median(c19$groceries), 2)),
                   as.character(round(median(c19$parks), 2)),
                   as.character(round(median(c19$transit), 2)),
                   as.character(round(median(c19$work), 2)),
                   as.character(round(median(c19$residential), 2))),
           max = c(as.character(max(c19$new_cases)),
                   as.character(max(c19$date)),
                   as.character(round(max(c19$retail), 2)),
                   as.character(round(max(c19$groceries), 2)),
                   as.character(round(max(c19$parks), 2)),
                   as.character(round(max(c19$transit), 2)),
                   as.character(round(max(c19$work), 2)),
                   as.character(round(max(c19$residential), 2))),
           sd = c(as.character(round(sd(c19$new_cases), 2)),
                   as.character(""),
                   as.character(round(sd(c19$retail), 2)),
                   as.character(round(sd(c19$groceries), 2)),
                   as.character(round(sd(c19$parks), 2)),
                   as.character(round(sd(c19$transit), 2)),
                   as.character(round(sd(c19$work), 2)),
                   as.character(round(sd(c19$residential), 2)))) %>%
  kable("latex",
        booktabs = TRUE,
        caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days. Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated. This variable (log-transformed after adding a small constant) is paired to the corresponding lagged moving average of the mobility indicators. The log-transformation is useful to avoid negative values of daily new cases when making predictions. Analysis is based on correlation analysis, multivariate regression, and data visualization. Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set. Temporal coverage is from February 27 to April 26, 2020. The maximum number of new daily cases during this period is 12,126.

Findings
===================

Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other. This is not surprising: it is well-known that mobility involves time-use trade-offs: the more there is of one kind of mobility, the less time there is available for any other. That said, the strongest correlation with the outcome of interest is for residential-based mobility. To avoid multicollinearity, parks-related mobility is selected as a covariate since it has the lowest correlation with residential-based mobility.

```{r check-correlations, echo=FALSE}
#Check the correlation between the mobility indicators and incidence:
c19 %>% 
  select(log_new_cases, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

A regression model uses the log of new daily cases as the dependent variable. The covariates are parks- and residential-related mobility, which enter the regression in the form of a second order polynomial expansion. In addition, the date is introduced to account for the temporal trend of the pandemic. Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there. The results appear in Table \ref{tab:model-results}. The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.

There is an overall temporal trend that indicates a growing number of cases over time, but at a decreasing rate (see negative sign of date^2). New York has on average more new daily cases than the rest of the states, but this has also declined over time (see negative sign of NY x date). Visualization is the most effective way to understand the trend according to the mobility indicators and date. Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states are considering relaxing stay-at-home orders. On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat. This changes by April 5, when large increases in residential-based mobility are seen, along with greater variations in parks-based mobility. The prediction surface indicates an expectation of a greater number of new daily cases as both classes of mobility increase. On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions. In general, higher levels of mobility tend to be associates with a higher number of new daily cases.

```{r model-estimation, include=FALSE}
#This is the regression model (forcing the regression through the intercept):
mod1 <- lm(log_new_cases ~  0 +
             day +
             day2 +
             parks2 + 
             parks + 
             parksXresidential +
             residential +
             residential2 +
             parksXday + 
             #parksXday2 +
             residentialXday +
             residentialXday2 +
             NY + 
             NYxday, 
           data = c19 #%>% filter(date >= "2020-04-01")
           )
summ(mod1)
```

```{r model-results, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x residential",
                        "residential",
                        "residential^2",
                        "parks x date",
                        "residential x date",
                        "residential x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

```{r summary-of-variables-for-prediction-grid, include=FALSE}
c19 %>% 
  #filter(day == 59) %>% 
  select(date, day, parks, residential) %>%
  summary()
```

```{r create-prediction-grid, include=FALSE}
new.data = expand.grid(parks = seq(0.36, 1.798, by = 0.01),
                       residential = seq(0.9736, 1.26, by = 0.01),
                       day = seq(37, 96, 1),
                       NY = c(0, 1)) %>%
  mutate(parks2 = parks^2,
         residential2 = residential^2,
         parksXresidential = parks * residential,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks2 * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NYxday = NY * day)

new.data <- data.frame(new.data, log_new_cases = predict(mod1, newdata = new.data))
```

```{r check-dates-for-examples, include=FALSE}
c19 %>% filter(date == "2020-04-20" | date == "2020-03-21" | date == "2020-04-05" ) %>% select(date, day) %>% distinct()
```

```{r check-mobility-indicators-on-2020-03-21, include=FALSE}
c19 %>% filter(date == "2020-03-21") %>% select(day, parks, residential) %>% summary()
```

```{r plot-03-21, include=FALSE, fig.cap="\\label{fig:plot-03-21}Prediction surface at various levels of the mobility indicators on March 21, 2020"}
plot_03_21 <- ggplot(data = new.data %>% filter(day == 60, parks >= 0.8527 & parks <= 1.5555, residential >= 1.025 & residential <= 1.09),
       aes(x = parks,
           y = residential)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "white") +
  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
  geom_point(data = c19 %>% filter(day == 60, state != "New York" & 
                                     state != "District of Columbia") %>% 
               select(parks, residential)) +
  geom_text(data = c19 %>% 
              filter(day == 60,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, residential, state), 
            aes(label = state),
            nudge_y = -0.01) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("March 21, 2020")
```

```{r check-mobility-indicators-on-2020-04-05, include=FALSE}
c19 %>% filter(date == "2020-04-05") %>% select(day, parks, residential) %>% summary()
```

```{r plot-04-05, include=FALSE, fig.cap="\\label{fig:plot-04-05}Prediction surface at various levels of the mobility indicators on April 05, 2020"}
plot_04_05 <- ggplot(data = new.data %>% filter(day == 75, parks >= 0.4545 & parks <= 1.6009, residential >= 1.122 & residential <= 1.253),
       aes(x = parks,
           y = residential)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "white") +
  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
  geom_point(data = c19 %>% filter(day == 75, state != "New York" & 
                                     state != "District of Columbia") %>% 
               select(parks, residential)) +
  geom_text(data = c19 %>% 
              filter(day == 75,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, residential, state), 
            aes(label = state),
            nudge_y = -0.01) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("April 05, 2020")
```

```{r check-mobility-indicators-on-2020-04-20, include=FALSE}
c19 %>% filter(date == "2020-04-20") %>% select(day, parks, residential) %>% summary()
```

```{r plot-04-20, include=FALSE, fig.cap="\\label{fig:plot-04-20}Prediction surface at various levels of the mobility indicators on April 20, 2020"}
plot_04_20 <- ggplot(data = new.data %>% filter(day == 90, parks >= 0.36 & parks <= 1.6309, residential >= 1.116 & residential <= 1.260),
       aes(x = parks,
           y = residential)) +
  geom_tile(aes(fill = log_new_cases)) + 
  geom_contour(aes(z = log_new_cases), color = "white") +
  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
  geom_point(data = c19 %>% filter(day == 90, state != "New York" & 
                                     state != "District of Columbia") %>% 
               select(parks, residential)) +
  geom_text(data = c19 %>% 
              filter(day == 90,
                     state == "Arkansas" |
                       state == "Mississippi" |
                       state == "Oklahoma" |
                       state == "Idaho" |
                       state == "New Jersey" |
                       state == "Michigan" |
                       state == "California") %>%
              select(parks, residential, state), 
            aes(label = state),
            nudge_y = -0.01) +
  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
  theme_light() +
  ggtitle("April 20, 2020")
```

```{r interactive-plot, eval=FALSE,include=FALSE}
plot_ly(data = c19 %>% filter(day == 90) %>% select(state, retail, parks), x = ~ parks, y = ~ retail, text = ~state)
```

```{r prediction-plots, fig.height=9, echo=FALSE, fig.cap="\\label{fig:prediction-plots}Prediction surfaces at three points during the pandemic according to the model; the dots are scatterplots of the park- and residential-mobility indicators of the states on that date."}
grid.arrange(plot_03_21, plot_04_05, plot_04_20, nrow = 3)
```

The results suggest the potential of GCMR to investigate the spread of COVID-19, but also point at some limitations. The baseline level is not defined in a metric that is amenable to policy development (Google defines "mobility" as an aggregated score of visits and length of stay at places.) For example, it is not clear precisely what is residential mobility: is it visits with friends and relatives, or mobility in the vicinity of the place of residency? Without a clearer understanding of these variables, their use can suggest trends, but their potential for applied policy analysis appears to be more limited.

References {#references .unnumbered}
==========

### 2020-05-04 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.tex
Initial commit

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Transportation Findings} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Using Google Community Mobility Reports to investigate the growth of
COVID-19 in the United States}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, Hamilton,
ON, L8S 4K1, Canada}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  In 2020 Google released a set of Community Mobility Reports (GCMR).
  These reports are based on the company's location-tracking capabilities
  and measure changes in mobility with respect to a baseline. This novel
  source of data offers an opportunity to investigate the way mobility
  potentially correlates with transmission of COVID-19. Using data from
  the New York Times on COVID-19 cases and GCMR, this paper presents an
  analysis of mobility levels and new daily reported cases of COVID-19 by
  state in the US. The results provide insights about the utility and
  interpretability of GCMR for COVID-19 research and decision-making.
  \end{abstract}

 \end{frontmatter}

\hypertarget{research-questions-and-hypotheses}{%
\section{Research Questions and
Hypotheses}\label{research-questions-and-hypotheses}}

The main policy tool to control the spread of the COVID-19 pandemic has
been stay-at-home orders. Concurrently, numerous efforts exist to track
the progress and the impact of the pandemic, and new sources of data
include the recently-released Google Community Mobility Reports
(GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as
The New York Times repository of COVID-19
data\footnote{https://github.com/nytimes/covid-19-data}. These two open
data sets offer novel opportunities to investigate in quasi-real time
the relationship between mobility patterns and transmission of COVID-19.

This paper investigates the potential of Google Community Mobility
Reports to asses the impact of mobility on COVID-19. The following
questions are posed:

\begin{itemize}
\tightlist
\item
  Do changes in mobility according to GCMR correlate with the
  transmission of COVID-19?
\item
  And if so, what do we learn about mobility and the spread of the
  disease?
\end{itemize}

The source document for this paper is an \texttt{R} markdown document
available from the following repository:

https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US

\hypertarget{methods-and-data}{%
\section{Methods and Data}\label{methods-and-data}}

GCMR use aggregated and anonymized data to chart changes in mobility
with respect to different classes of places, namely retail and
recreation, groceries and pharmacies, parks, transit stations,
workplaces, and residential. Mobility indicators for each of these
places are reported as percentage changes from a baseline level, which
corresponds to the median value of mobility of identical days of the
week during the period between January 3 and Feb 6, 2020. Covid-19 data
is compiled by The New York Times based on reports from state and local
health agencies.

For analysis, all mobility indicators are centered so that the value of
1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
change. The incubation time of the disease has been estimated to be
between 2 and 12 days (95\% interval) by Lauer et al.~(2020). Given
this, it is to be expected that any changes in mobility will have a
lagged effect on the discovery of new cases. For this reason, lagged
moving averages of the mobility indicators are calculated. Furthermore,
it is possible that mobility and reports of new cases of COVID-19 are
endogenous, if the public adjust their mobility according to reports of
the incidence. Therefore, in addition to being consistent with an
incubation period, use of lagged indicators also helps to break this
potential endogeneity.

\begin{table}[!h]

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics}Descriptive statistics of the data set}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{llllll}
\toprule
Variable & Definition & min & median & max & sd\\
\midrule
\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 77.5 & 12126 & 1059.81\\
date & Date & 2020-02-27 & 2020-04-01 & 2020-04-26 & \\
\rowcolor{gray!6}  retail & Retail and Recreation & 0.34 & 0.66 & 1.16 & 0.22\\
groceries & Groceries and Pharmacies & 0.66 & 0.95 & 1.26 & 0.13\\
\rowcolor{gray!6}  parks & Parks & 0.36 & 1.15 & 1.8 & 0.24\\
\addlinespace
transit & Transit Stations & 0.24 & 0.73 & 1.14 & 0.24\\
\rowcolor{gray!6}  work & Workplaces & 0.34 & 0.65 & 1.07 & 0.2\\
residential & Residential & 0.97 & 1.14 & 1.27 & 0.08\\
\bottomrule
\multicolumn{6}{l}{\textit{Note: }}\\
\multicolumn{6}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

The lagged indicators are calculated as the mean of the mobility
indicator using the values from date-minus-12-days to date-minus-2-days.
Furthermore, using the cumulative number of reported COVID-19 cases, the
total number of new daily cases is calculated. This variable
(log-transformed after adding a small constant) is paired to the
corresponding lagged moving average of the mobility indicators. The
log-transformation is useful to avoid negative values of daily new cases
when making predictions. Analysis is based on correlation analysis,
multivariate regression, and data visualization. Table
\ref{tab:descriptive-statistics} shows the descriptive statistics of the
data set. Temporal coverage is from February 27 to April 26, 2020. The
maximum number of new daily cases during this period is 12,126.

\hypertarget{findings}{%
\section{Findings}\label{findings}}

Table \ref{tab:correlation-analysis} shows that the mobility indicators
are highly correlated with each other. This is not surprising: it is
well-known that mobility involves time-use trade-offs: the more there is
of one kind of mobility, the less time there is available for any other.
That said, the strongest correlation with the outcome of interest is for
residential-based mobility. To avoid multicollinearity, parks-related
mobility is selected as a covariate since it has the lowest correlation
with residential-based mobility.

\begin{table}[!h]

\caption{\label{tab:check-correlations}\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrr}
\toprule
  & log\_new\_cases & retail & groceries & parks & transit & work & residential\\
\midrule
\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.47 & -0.30 & -0.68 & -0.69 & 0.71\\
retail & -0.68 & 1.00 & 0.86 & 0.44 & 0.93 & 0.98 & -0.98\\
\rowcolor{gray!6}  groceries & -0.47 & 0.86 & 1.00 & 0.46 & 0.87 & 0.87 & -0.86\\
parks & -0.30 & 0.44 & 0.46 & 1.00 & 0.52 & 0.44 & -0.45\\
\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.52 & 1.00 & 0.94 & -0.95\\
\addlinespace
work & -0.69 & 0.98 & 0.87 & 0.44 & 0.94 & 1.00 & -0.99\\
\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.86 & -0.45 & -0.95 & -0.99 & 1.00\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{All mobility indicators are lagged 11-day moving averages}\\
\end{tabular}}
\end{table}

A regression model uses the log of new daily cases as the dependent
variable. The covariates are parks- and residential-related mobility,
which enter the regression in the form of a second order polynomial
expansion. In addition, the date is introduced to account for the
temporal trend of the pandemic. Finally, an indicator variable for the
state of New York and an interaction with the date are used to
distinguish the unusually high incidence of the disease there. The
results appear in Table \ref{tab:model-results}. The model provides a
good fit to the data and all variables reported are significant at
\(p<0.05\) or better.

There is an overall temporal trend that indicates a growing number of
cases over time, but at a decreasing rate (see negative sign of
date\^{}2). New York has on average more new daily cases than the rest
of the states, but this has also declined over time (see negative sign
of NY x date). Visualization is the most effective way to understand the
trend according to the mobility indicators and date. Figure
\ref{fig:prediction-plots} shows the prediction surfaces on three
different dates: March 21, when the first states began implementing
stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
more days later, and at a time when some states are considering relaxing
stay-at-home orders. On March 21 there were still only minor departures
from baseline mobility (recall that these are temporally lagged); the
prediction surface at this point is relatively flat. This changes by
April 5, when large increases in residential-based mobility are seen,
along with greater variations in parks-based mobility. The prediction
surface indicates an expectation of a greater number of new daily cases
as both classes of mobility increase. On the last date examined, April
20, the trend becomes more steep for park-based mobility, even as this
indicator continues to display large variations from the baseline in
both directions. In general, higher levels of mobility tend to be
associates with a higher number of new daily cases.

\begin{table}[!h]

\caption{\label{tab:model-results}\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).}
\centering
\begin{tabular}[t]{lcc}
\toprule
Variable & Coefficient Estimate & p-value\\
\midrule
\rowcolor{gray!6}  date & 7.9619 & <0.0001\\
date\textasciicircum{}2 & -0.0411 & <0.0001\\
\rowcolor{gray!6}  parks\textasciicircum{}2 & 1.4017 & 0.0251\\
parks & -37.2492 & <0.0001\\
\rowcolor{gray!6}  parks x residential & 39.0902 & <0.0001\\
\addlinespace
residential & -286.7568 & <0.0001\\
\rowcolor{gray!6}  residential\textasciicircum{}2 & 249.9901 & <0.0001\\
parks x date & -0.1412 & <0.0001\\
\rowcolor{gray!6}  residential x date & -6.9180 & <0.0001\\
residential x date\textasciicircum{}2 & 0.0365 & <0.0001\\
\addlinespace
\rowcolor{gray!6}  NY & 6.9864 & <0.0001\\
NY x date & -0.0557 & 0.0022\\
\bottomrule
\multicolumn{3}{l}{\textit{Note: }}\\
\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.821}\\
\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.82}\\
\multicolumn{3}{l}{Standard Error $\sigma$= 2.102}\\
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{Covid-19-Google-CMR-US_files/figure-latex/prediction-plots-1.pdf}
\caption{\label{fig:prediction-plots}Prediction surfaces at three points
during the pandemic according to the model; the dots are scatterplots of
the park- and residential-mobility indicators of the states on that
date.}
\end{figure}

The results suggest the potential of GCMR to investigate the spread of
COVID-19, but also point at some limitations. The baseline level is not
defined in a metric that is amenable to policy development (Google
defines ``mobility'' as an aggregated score of visits and length of stay
at places.) For example, it is not clear precisely what is residential
mobility: is it visits with friends and relatives, or mobility in the
vicinity of the place of residency? Without a clearer understanding of
these variables, their use can suggest trends, but their potential for
applied policy analysis appears to be more limited.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Lauer2020incubation}{}%
Lauer, S.A., Grantz, K.H., Bi, Q., Jones, F.K., Zheng, Q., Meredith,
H.R., Azman, A.S., Reich, N.G., Lessler, J., 2020. The incubation period
of coronavirus disease 2019 (covid-19) from publicly reported confirmed
cases: Estimation and application. Annals of Internal Medicine.
doi:\href{https://doi.org/10.7326/m20-0504}{10.7326/m20-0504}

\end{document}

### 2020-05-04 README.Rmd
Created README.Rmd

--- 2020-05-04 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
@@
 ---
-title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
-author:
-  - name: Antonio Paez
-    email: paezha@mcmaster.ca
-    affiliation: McMaster University
-    footnote: 1
-address:
-  - code: McMaster University
-    address: School of Geography and Earth Sciences, McMaster University, Hamilton, ON, L8S 4K1, Canada
-footnote:
-  - code: 1
-    text: "Corresponding Author"
-abstract: |
+output: github_document
+always_allow_html: true
+bibliography: References.bib
+---
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>"
+)
+```
+
+# Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
+
+<!-- badges: start -->
+<!-- badges: end -->
+
+## Abstract
+
 In 2020 Google released a set of Community Mobility Reports (GCMR).
@@
 
-journal: "Transportation Findings"
-date: "`r Sys.Date()`"
-bibliography: References.bib
-#linenumbers: true
-#numbersections: true
-csl: elsevier-harvard.csl
-output: rticles::elsevier_article
----
+## Keywords
+
+- COVID-19
+- Google Community Mobility Reports
+- Mobility
+- Modelling
+- Regression analysis
+- Policy
 
@@
 The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders.
-Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}.
+Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports [GCMR](https://www.google.com/covid19/mobility/), as well as The New York Times repository of [COVID-19 data](https://github.com/nytimes/covid-19-data).
 These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.
@@
 
-https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US
+https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/tree/fcd055878020bf567cd4deabb16c8056189eb45c/Covid-19-Google-CMR-US
 
@@
                    as.character(round(sd(c19$residential), 2)))) %>%
-  kable("latex",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
   cor(use = "pairwise.complete.obs") %>%
-  kable("latex",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
         caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
-  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
+  #kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
   mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
-  kable("latex",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
         col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
-  kable_styling(latex_options = c("striped", "hold_position")) %>%
+  #kable_styling(latex_options = c("striped", "hold_position")) %>%
   footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),

### 2020-05-11 Covid-19-Google-CMR-US/Response-to-Reviewers-v2.Rmd
Second version of paper after first round of anonymous reviews.

---
title: "Response to Reviewers"
author: "Antonio Paez"
date: "5/11/2020"
output: pdf_document
---

I would like to thank two anonymous reviewers for their comments and suggestions for revisions. In this letter I respond to those comments and explain how the paper was revised in response.

\newpage

# Reviewer 1

This is a well-written and timely paper. I only have some comments regarding the interpretation. So Table 3 indicates that park-related mobility (I assume this refers to walking and cycling in a park? this could be specified) results in a new/more cases of Covid 19, while residential mobility (mobility in own neighborhood?) results in a lower amount of Covid 19 cases? I think this could be stated somewhat more clear/ easier to understand in the paper.

>*Thank you for your positive assessment of the paper, and comments for improving it. The parks mobility index refers to mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens. Following a recommendation from Reviewer 2 I have replaced "residential mobility" with "work-related mobility", which is easier to interpret. Hopefully in the revision the results are more clearly stated.*

I would also suggest adding a sentence (before of after table 3) like: ‘These results indicate that traveling outside the residential neighborhood increases the chance of getting infected by COVID-19, and therefore aligns with existing social distancing rules’ (or a similar sentence refering to policy implication)

>*Thank you. I added a sentence regarding the policy implications of the analysis.*

\newpage

# Reviewer 2

The question the paper deals with is timely and interesting. However, there are a number of issues. First, the intent of the stay-at-home orders is to reduce transmission from what it would have been otherwise. Cases are going up regradless, but if the policy works, there would be fewer cases. Without a counterfactual of what the number of infections would have been, it is impossible to find meaning from the model. It is the reductions that should be a response variable to mobility changes.

>*That is correct. In lieu of a counterfactual, what we have is differences in compliance with the policy in different states, which are reflected statistically in different rates in the growth of cases. See for example the table below. On March 21, New Jersey had similary mobility with respect to parks and work as Idaho. However, these two states had a different starting point for the trajectory of the pandemic - which is reflected in the number of new cases on that date (both actual and estimated by the model). By April 20, New Jersey had substantially reduced park-related mobility, whereas Idaho's was even higher. The estimated and actual number of new cases grew in the intervening period; however, New Jersey's growth in new cases (from March 21 to April 20) was only 894%, whereas Idaho's was 1030%.*

State | Date | Parks | Work | Estimated New Cases | Actual New Cases |
-| - | - | - | - | - |
New Jersey | March 21 | 1.45 | 0.845 | 441.42 | 440 |
New Jersey | April 20 | 0.728 | 0.413 | 3498.19 | 4377 |
Idaho | March 21 | 1.48 | 0.939 | 23.10 | 23
Idaho | April 20 | 1.63 | 0.588 | 259.82 | 260

Second, the chosen mobility indicators (parks and residence) both increased from baseline in the period of analysis for most states (median > 1 for both variables). Neither captures the policy intent. Third, the residence mobility variable is not well defined as the author indicates in their conclusion. Other variables such as work, transit better capture the policy intent and are better defined.

>*I am grateful for these comments, which made me rethink the selection of variables. Here, it is important to make a distinction between the _intent_ of the policy, and _compliance_ with the policy. While the policy's intent has been to reduce out-of-home activities to limit the contact rate for the virus, compliance has been haphazard, with some states (such as New Jersey in the example above) recording sharper declines in mobility, and others (such as Idaho) where compliance has been more lax.*

>*The above said, I was encouraged by this comment to replace residential mobility (which I struggled with the interpretation of in any case) and use instead work-related mobility, while retaining park-related mobility. Why these two variables? First, as seen in Table 2, the correlations between the variables are very high, which means that after choosing any one, a second mobility index does not bring a lot of new information. The exception is parks, the only variable with R < 0.8 with any of the other mobility indicators. So partly, I chose to keep parks in the analysis for statistical reasons. Secondly, work and parks capture two dimensions of mobility, mandatory and discretionary travel, so conceptually they make sense too. And finally, these two mobility indicators are better defined than residential-based mobility.*

>*The results indicate that the number of new cases generally increases as a function of mobility, but the effect is not linear, given the interactions between work-related and parks-related mobility.*

Finally, even though the author states that “higher levels of mobility tend to be associated with higher number of cases”, that is not what the estimated parameters show within their range in the data for both residence and parks. For example, taking the quadratic function for residences, within the range of 0.34-1.16, infections decline as mobility increases - the opposite of what the policy is supposed to do. It seems that the declines in the included figures are driven by the date variable and interactions with it.

>*There is indeed a strong temporal trend, which is captured by the date and interaction of date with the mobility indicators. The coefficients of the model indicate that the number of new cases increases with park-related mobility, and also increases with work related mobility. The interaction term is negative, which captures the trade-off between these two forms of mobility. As seen in the figures, the trend is stronger in the dimension of parks-related mobility: as trade-offs go, gains due to lower work-related mobility can be more than offset by parks-related mobility.*

### 2020-05-11 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
Updated link to repository
Second version of paper after first round of anonymous reviews.

@@
 These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline.
-This novel source of data offers an opportunity to investigate the way mobility potentially correlates with transmission of COVID-19.
+This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19.
 Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US.
@@
 output: rticles::elsevier_article
+always_allow_html: true
 ---
@@
 c19 <- c19 %>%
-  mutate(day = yday(date) - min(yday(date)))
+  mutate(day = yday(date) - yday("2020-04-05"))
 ```
@@
          work = Mean_work_lag11,
+         work2 = work^2,
          residential = Mean_residential_lag11,
@@
          parksXresidential = parks * residential,
+         parksXwork = parks * work,
          day2 = day^2,
@@
          parksXday2 = parks * day2,
+         workXday = work * day,
+         workXday2 = work * day2,
          residentialXday = residential * day,
@@
 
-The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders.
-Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}.
+The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance.
+Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic.
+As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}.
 These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.
@@
 
-The source document for this paper is an `R` markdown document available from the following repository:
-
-https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US
+This paper is a reproducible research document.
+The source is an `R` markdown file available in a public repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.
 
@@
 
-GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places, namely retail and recreation, groceries and pharmacies, parks, transit stations, workplaces, and residential.
-Mobility indicators for each of these places are reported as percentage changes from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020.
+GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}).
+Mobility indicators are calculated based on the frequency and length of visits.
+The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020.
 Covid-19 data is compiled by The New York Times based on reports from state and local health agencies.
-
-For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change.
-The incubation time of the disease has been estimated to be between 2 and 12 days (95% interval) by Lauer et al.
-[-@Lauer2020incubation].
-Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases.
-For this reason, lagged moving averages of the mobility indicators are calculated.
-Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence.
-Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.
 
@@
                         "Date",
-                        "Retail and Recreation",
-                        "Groceries and Pharmacies",
-                        "Parks",
-                        "Transit Stations",
-                        "Workplaces",
-                        "Residential"),
+                        "Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters",
+                        "Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies",
+                        "Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens",
+                        "Mobility trends for places like public transport hubs such as subway, bus, and train stations",
+                        "Mobility trends for places of work",
+                        "Mobility trends for places of residence"),
            min = c(as.character(min(c19$new_cases)),
@@
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
-  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
+  column_spec(2, width = "12em") %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
 
+For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change.
+The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation].
+Given this, it is to be expected that any changes in mobility will have a lagged effect (if any) on the discovery of new cases.
+For this reason, lagged moving averages of the mobility indicators are calculated.
+Furthermore, it is possible that mobility and reports of new cases of COVID-19 are endogenous, if the public adjust their mobility according to reports of the incidence.
+Therefore, in addition to being consistent with an incubation period, use of lagged indicators also helps to break this potential endogeneity.
+
 The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days.
 Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated.
-This variable (log-transformed after adding a small constant) is paired to the corresponding lagged moving average of the mobility indicators.
+This variable (log-transformed after adding a small constant) is paired with the corresponding lagged moving average of the mobility indicators.
 The log-transformation is useful to avoid negative values of daily new cases when making predictions.
+Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set.
 Analysis is based on correlation analysis, multivariate regression, and data visualization.
-Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set.
-Temporal coverage is from February 27 to April 26, 2020.
-The maximum number of new daily cases during this period is 12,126.
 
@@
 Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other.
-This is not surprising: it is well-known that mobility involves time-use trade-offs: the more there is of one kind of mobility, the less time there is available for any other.
-That said, the strongest correlation with the outcome of interest is for residential-based mobility.
-To avoid multicollinearity, parks-related mobility is selected as a covariate since it has the lowest correlation with residential-based mobility.
+Two variables are selected for multivariate analysis: parks- and work-related mobility.
+Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables.
+Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.
 
@@
         caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
-  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
 
-A regression model uses the log of new daily cases as the dependent variable.
-The covariates are parks- and residential-related mobility, which enter the regression in the form of a second order polynomial expansion.
-In addition, the date is introduced to account for the temporal trend of the pandemic.
+A regression model is estimated with the log of new daily cases as the dependent variable.
+The covariates enter the regression in the form of a second order polynomial expansion.
+In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic.
 Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there.
@@
 
-There is an overall temporal trend that indicates a growing number of cases over time, but at a decreasing rate (see negative sign of date^2).
-New York has on average more new daily cases than the rest of the states, but this has also declined over time (see negative sign of NY x date).
-Visualization is the most effective way to understand the trend according to the mobility indicators and date.
-Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states are considering relaxing stay-at-home orders.
-On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
-This changes by April 5, when large increases in residential-based mobility are seen, along with greater variations in parks-based mobility.
-The prediction surface indicates an expectation of a greater number of new daily cases as both classes of mobility increase.
-On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
-In general, higher levels of mobility tend to be associates with a higher number of new daily cases.
+There is an overall temporal trend that indicates a growing number of cases over time, at an accelerating rate (see positive sign of date^2).
+Mobility related to parks and to work both tend to increase the number of new cases; the negative sign for their interaction is indicative of the trade-offs between these two forms of mobility and their impact on new cases.
+The influence of parks-related mobility was relatively weaker early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term).
+The opposite happens with work-related mobility, which started with a greater effect (positive sign of work x date term), but whose impact has declined over time (negative sign of work x date^2 term) New York has on average more new daily cases than the rest of the states, but this has declined over time (see negative sign of NY x date).
 
@@
              parks + 
-             parksXresidential +
-             residential +
-             residential2 +
+             parksXwork +
+             work +
+             work2 +
              parksXday + 
-             #parksXday2 +
-             residentialXday +
-             residentialXday2 +
+             parksXday2 +
+             workXday +
+             workXday2 +
              NY + 
@@
                         "parks",
-                        "parks x residential",
-                        "residential",
-                        "residential^2",
+                        "parks x work",
+                        "work",
+                        "work^2",
                         "parks x date",
-                        "residential x date",
-                        "residential x date^2",
+                        "parks x date^2",
+                        "work x date",
+                        "work x date^2",
                         "NY",
@@
            pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
-  mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
+  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
   kable("latex",
@@
         col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
-  kable_styling(latex_options = c("striped", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
   footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
@@
 ```
+
+Visualization is the most effective way to understand the trend according to the mobility indicators and date.
+Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states started to consider relaxing stay-at-home orders.
+
+On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
+This changes by April 5, when work-based mobility has declined substantially.
+Although every state registers lower work-based mobility, there are large variations in parks-based mobility, with some states seeing increases of up to 60% for this class of mobility.
+
+The prediction surface indicates an expectation of a greater number of new daily cases as either class of mobility increases, but the effect is not linear.
+On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
+The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the emergence of new cases.
 
@@
 ```{r create-prediction-grid, include=FALSE}
-new.data = expand.grid(parks = seq(0.36, 1.798, by = 0.01),
-                       residential = seq(0.9736, 1.26, by = 0.01),
-                       day = seq(37, 96, 1),
+new.data = expand.grid(parks = seq(0.36, 1.9, by = 0.01),
+                       work = seq(0.33, 1.08, by = 0.01),
+                       day = seq(-15, 15, 1),
                        NY = c(0, 1)) %>%
   mutate(parks2 = parks^2,
-         residential2 = residential^2,
-         parksXresidential = parks * residential,
+         work2 = work^2,
+         parksXwork = parks * work,
          day2 = day^2,
@@
          parksXday2 = parks2 * day2,
-         residentialXday = residential * day,
-         residentialXday2 = residential * day2,
+         workXday = work * day,
+         workXday2 = work * day2,
          NYxday = NY * day)
@@
 ```{r check-mobility-indicators-on-2020-03-21, include=FALSE}
-c19 %>% filter(date == "2020-03-21") %>% select(day, parks, residential) %>% summary()
+c19 %>% filter(date == "2020-03-21") %>% select(day, parks, work) %>% summary()
 ```
@@
 ```{r plot-03-21, include=FALSE, fig.cap="\\label{fig:plot-03-21}Prediction surface at various levels of the mobility indicators on March 21, 2020"}
-plot_03_21 <- ggplot(data = new.data %>% filter(day == 60, parks >= 0.8527 & parks <= 1.5555, residential >= 1.025 & residential <= 1.09),
+plot_03_21 <- ggplot(data = new.data %>% filter(day == -15, parks >= 0.8527 & parks <= 1.5555, work >= 0.7627 & work <= 0.9409),
        aes(x = parks,
-           y = residential)) +
+           y = work)) +
   geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "white") +
-  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
-  geom_point(data = c19 %>% filter(day == 60, state != "New York" & 
-                                     state != "District of Columbia") %>% 
-               select(parks, residential)) +
+  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.43, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_point(data = c19 %>% filter(day == -15, state != "New York") %>% 
+               select(parks, work)) +
   geom_text(data = c19 %>% 
-              filter(day == 60,
+              filter(day == -15,
                      state == "Arkansas" |
@@
                        state == "California") %>%
-              select(parks, residential, state), 
+              select(parks, work, state), 
             aes(label = state),
             nudge_y = -0.01) +
+#  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
   scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
@@
 ```{r check-mobility-indicators-on-2020-04-05, include=FALSE}
-c19 %>% filter(date == "2020-04-05") %>% select(day, parks, residential) %>% summary()
+c19 %>% filter(date == "2020-04-05") %>% select(day, parks, work) %>% summary()
 ```
@@
 ```{r plot-04-05, include=FALSE, fig.cap="\\label{fig:plot-04-05}Prediction surface at various levels of the mobility indicators on April 05, 2020"}
-plot_04_05 <- ggplot(data = new.data %>% filter(day == 75, parks >= 0.4545 & parks <= 1.6009, residential >= 1.122 & residential <= 1.253),
+plot_04_05 <- ggplot(data = new.data %>% filter(day == 0, parks >= 0.4545 & parks <= 1.6009, work >= 0.3673 & work <= 0.6955),
        aes(x = parks,
-           y = residential)) +
+           y = work)) +
   geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "white") +
-  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
-  geom_point(data = c19 %>% filter(day == 75, state != "New York" & 
-                                     state != "District of Columbia") %>% 
-               select(parks, residential)) +
+  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.162, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_point(data = c19 %>% filter(day == 0, state != "New York") %>% 
+               select(parks, work)) +
   geom_text(data = c19 %>% 
-              filter(day == 75,
+              filter(day == 0,
                      state == "Arkansas" |
@@
                        state == "California") %>%
-              select(parks, residential, state), 
+              select(parks, work, state), 
             aes(label = state),
             nudge_y = -0.01) +
+  #scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
   scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
@@
 ```{r check-mobility-indicators-on-2020-04-20, include=FALSE}
-c19 %>% filter(date == "2020-04-20") %>% select(day, parks, residential) %>% summary()
+c19 %>% filter(date == "2020-04-20") %>% select(day, parks, work) %>% summary()
 ```
@@
 ```{r plot-04-20, include=FALSE, fig.cap="\\label{fig:plot-04-20}Prediction surface at various levels of the mobility indicators on April 20, 2020"}
-plot_04_20 <- ggplot(data = new.data %>% filter(day == 90, parks >= 0.36 & parks <= 1.6309, residential >= 1.116 & residential <= 1.260),
+plot_04_20 <- ggplot(data = new.data %>% filter(day == 15, parks >= 0.36 & parks <= 1.6309, work >= 0.3582 & work <= 0.6518),
        aes(x = parks,
-           y = residential)) +
+           y = work)) +
   geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "white") +
-  #geom_text_contour(aes(z = incidence), stroke = 0.1, size = 3.5) +
-  geom_point(data = c19 %>% filter(day == 90, state != "New York" & 
-                                     state != "District of Columbia") %>% 
-               select(parks, residential)) +
+  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.18, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_point(data = c19 %>% filter(day == 15, state != "New York") %>% 
+               select(parks, work)) +
   geom_text(data = c19 %>% 
-              filter(day == 90,
+              filter(day == 15,
                      state == "Arkansas" |
@@
 
-```{r interactive-plot, eval=FALSE,include=FALSE}
-plot_ly(data = c19 %>% filter(day == 90) %>% select(state, retail, parks), x = ~ parks, y = ~ retail, text = ~state)
-```
-
-```{r prediction-plots, fig.height=9, echo=FALSE, fig.cap="\\label{fig:prediction-plots}Prediction surfaces at three points during the pandemic according to the model; the dots are scatterplots of the park- and residential-mobility indicators of the states on that date."}
+```{r prediction-plots, fig.height=9, echo=FALSE, fig.cap="\\label{fig:prediction-plots}Prediction surfaces at three points during the pandemic according to the model; the dots are a scatterplot of the parks- and work-related mobility indicators of the states on that date; the white dashed line is the fold of the saddle."}
 grid.arrange(plot_03_21, plot_04_05, plot_04_20, nrow = 3)
@@
 
-The results suggest the potential of GCMR to investigate the spread of COVID-19, but also point at some limitations.
-The baseline level is not defined in a metric that is amenable to policy development (Google defines "mobility" as an aggregated score of visits and length of stay at places.) For example, it is not clear precisely what is residential mobility: is it visits with friends and relatives, or mobility in the vicinity of the place of residency?
-Without a clearer understanding of these variables, their use can suggest trends, but their potential for applied policy analysis appears to be more limited.
+```{r plot-3d-surface, include=FALSE}
+# Choose day for the plot: April 20 is day == 15, April 5 is day == 0, March 21 is day == -15
+plotly.data <- new.data %>% filter(day == 15) 
+parks <- seq(min(plotly.data$parks), max(plotly.data$parks), by = 0.01)
+work <- seq(min(plotly.data$work), max(plotly.data$work), by = 0.01)
+
+log_new_cases <- xtabs(log_new_cases ~ work + parks, plotly.data)
+
+plot_ly(z = ~log_new_cases, x = ~parks, y = ~work) %>% 
+  add_surface() %>%
+  layout(scene = list(
+      xaxis = list(title = "parks"),
+      yaxis = list(title = "work"),
+      zaxis = list(title = "log(New_Cases)"),
+      aspectmode='cube'))
+```
+
+The results suggest that over time the benefits of reduced work-related mobility can be offset by parks-related mobility.
+For example, on March 21, New Jersey and Idaho had similar levels of park-related mobility.
+By April 20, New Jersey had substantially reduced park-related mobility, whereas Idaho's was even higher than in March.
+The estimated and actual number of new cases grew in the intervening period; however, New Jersey's growth in new cases (from March 21 to April 20) was only 894%, whereas Idaho's was 1030%.
+
+These results suggest the potential of GCMR to investigate the spread of COVID-19, but also point at some limitations.
+The baseline level is not defined in a metric that is amenable to policy development (e.g., person-km travelled).
+Without a clearer understanding of the absolute levels of these variables, their use can suggest trends, but their potential for applied policy analysis appears to be more limited.
 

### 2020-05-11 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.tex
Updated link to repository
Second version of paper after first round of anonymous reviews.

@@
   and measure changes in mobility with respect to a baseline. This novel
-  source of data offers an opportunity to investigate the way mobility
-  potentially correlates with transmission of COVID-19. Using data from
-  the New York Times on COVID-19 cases and GCMR, this paper presents an
-  analysis of mobility levels and new daily reported cases of COVID-19 by
-  state in the US. The results provide insights about the utility and
+  source of data offers an opportunity to investigate potential
+  correlations between mobility and transmission of COVID-19. Using data
+  from the New York Times on COVID-19 cases and GCMR, this paper presents
+  an analysis of mobility levels and new daily reported cases of COVID-19
+  by state in the US. The results provide insights about the utility and
   interpretability of GCMR for COVID-19 research and decision-making.
@@
 The main policy tool to control the spread of the COVID-19 pandemic has
-been stay-at-home orders. Concurrently, numerous efforts exist to track
-the progress and the impact of the pandemic, and new sources of data
-include the recently-released Google Community Mobility Reports
+been stay-at-home orders, which in the United States have been
+implemented on a state-by-state basis, with considerable variations in
+compliance. Concurrently, numerous initiatives have been developed to
+track the progress and the impact of the pandemic. As a result, there
+are new sources of data such as the recently-released Google Community
+Mobility Reports
 (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as
@@
 
-The source document for this paper is an \texttt{R} markdown document
-available from the following repository:
-
-https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US
+This paper is a reproducible research document. The source is an
+\texttt{R} markdown file available in a public
+repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.
 
@@
 GCMR use aggregated and anonymized data to chart changes in mobility
-with respect to different classes of places, namely retail and
-recreation, groceries and pharmacies, parks, transit stations,
-workplaces, and residential. Mobility indicators for each of these
-places are reported as percentage changes from a baseline level, which
-corresponds to the median value of mobility of identical days of the
-week during the period between January 3 and Feb 6, 2020. Covid-19 data
-is compiled by The New York Times based on reports from state and local
-health agencies.
+with respect to different classes of places (see Table
+\ref{tab:descriptive-statistics}). Mobility indicators are calculated
+based on the frequency and length of visits. The reports give percentage
+change from a baseline level, which corresponds to the median value of
+mobility of identical days of the week during the period between January
+3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based
+on reports from state and local health agencies.
 
-For analysis, all mobility indicators are centered so that the value of
-1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
-change. The incubation time of the disease has been estimated to be
-between 2 and 12 days (95\% interval) by Lauer et al.~(2020). Given
-this, it is to be expected that any changes in mobility will have a
-lagged effect on the discovery of new cases. For this reason, lagged
-moving averages of the mobility indicators are calculated. Furthermore,
-it is possible that mobility and reports of new cases of COVID-19 are
-endogenous, if the public adjust their mobility according to reports of
-the incidence. Therefore, in addition to being consistent with an
-incubation period, use of lagged indicators also helps to break this
-potential endogeneity.
-
-\begin{table}[!h]
+\begin{table}[H]
 
@@
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llllll}
+\fontsize{7}{9}\selectfont
+\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{12em}llll}
 \toprule
@@
 \midrule
-\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 77.5 & 12126 & 1059.81\\
-date & Date & 2020-02-27 & 2020-04-01 & 2020-04-26 & \\
-\rowcolor{gray!6}  retail & Retail and Recreation & 0.34 & 0.66 & 1.16 & 0.22\\
-groceries & Groceries and Pharmacies & 0.66 & 0.95 & 1.26 & 0.13\\
-\rowcolor{gray!6}  parks & Parks & 0.36 & 1.15 & 1.8 & 0.24\\
+\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 90 & 12312 & 1051.67\\
+date & Date & 2020-02-27 & 2020-04-04 & 2020-05-02 & \\
+\rowcolor{gray!6}  retail & Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters & 0.34 & 0.66 & 1.16 & 0.21\\
+groceries & Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies & 0.66 & 0.94 & 1.26 & 0.13\\
+\rowcolor{gray!6}  parks & Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens & 0.36 & 1.15 & 2.1 & 0.26\\
 \addlinespace
-transit & Transit Stations & 0.24 & 0.73 & 1.14 & 0.24\\
-\rowcolor{gray!6}  work & Workplaces & 0.34 & 0.65 & 1.07 & 0.2\\
-residential & Residential & 0.97 & 1.14 & 1.27 & 0.08\\
+transit & Mobility trends for places like public transport hubs such as subway, bus, and train stations & 0.24 & 0.72 & 1.14 & 0.23\\
+\rowcolor{gray!6}  work & Mobility trends for places of work & 0.34 & 0.63 & 1.07 & 0.2\\
+residential & Mobility trends for places of residence & 0.97 & 1.14 & 1.27 & 0.08\\
 \bottomrule
@@
 
+For analysis, all mobility indicators are centered so that the value of
+1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
+change. The incubation time of the disease is between 2 and 12 days
+(95\% interval; see Lauer et al., 2020). Given this, it is to be
+expected that any changes in mobility will have a lagged effect (if any)
+on the discovery of new cases. For this reason, lagged moving averages
+of the mobility indicators are calculated. Furthermore, it is possible
+that mobility and reports of new cases of COVID-19 are endogenous, if
+the public adjust their mobility according to reports of the incidence.
+Therefore, in addition to being consistent with an incubation period,
+use of lagged indicators also helps to break this potential endogeneity.
+
 The lagged indicators are calculated as the mean of the mobility
@@
 total number of new daily cases is calculated. This variable
-(log-transformed after adding a small constant) is paired to the
+(log-transformed after adding a small constant) is paired with the
 corresponding lagged moving average of the mobility indicators. The
 log-transformation is useful to avoid negative values of daily new cases
-when making predictions. Analysis is based on correlation analysis,
-multivariate regression, and data visualization. Table
-\ref{tab:descriptive-statistics} shows the descriptive statistics of the
-data set. Temporal coverage is from February 27 to April 26, 2020. The
-maximum number of new daily cases during this period is 12,126.
+when making predictions. Table \ref{tab:descriptive-statistics} shows
+the descriptive statistics of the data set. Analysis is based on
+correlation analysis, multivariate regression, and data visualization.
 
@@
 Table \ref{tab:correlation-analysis} shows that the mobility indicators
-are highly correlated with each other. This is not surprising: it is
-well-known that mobility involves time-use trade-offs: the more there is
-of one kind of mobility, the less time there is available for any other.
-That said, the strongest correlation with the outcome of interest is for
-residential-based mobility. To avoid multicollinearity, parks-related
-mobility is selected as a covariate since it has the lowest correlation
-with residential-based mobility.
+are highly correlated with each other. Two variables are selected for
+multivariate analysis: parks- and work-related mobility. Work has a high
+correlation with the outcome variable, and its correlation with parks is
+relatively weak, which increases the information content of the two
+variables. Furthermore, parks- and work-related mobility represent two
+dimensions of out-of-home activities: mandatory and discretionary
+travel.
 
-\begin{table}[!h]
+\begin{table}[H]
 
@@
 \resizebox{\linewidth}{!}{
+\fontsize{7}{9}\selectfont
 \begin{tabular}[t]{lrrrrrrr}
@@
 \midrule
-\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.47 & -0.30 & -0.68 & -0.69 & 0.71\\
-retail & -0.68 & 1.00 & 0.86 & 0.44 & 0.93 & 0.98 & -0.98\\
-\rowcolor{gray!6}  groceries & -0.47 & 0.86 & 1.00 & 0.46 & 0.87 & 0.87 & -0.86\\
-parks & -0.30 & 0.44 & 0.46 & 1.00 & 0.52 & 0.44 & -0.45\\
-\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.52 & 1.00 & 0.94 & -0.95\\
+\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.48 & -0.27 & -0.68 & -0.69 & 0.71\\
+retail & -0.68 & 1.00 & 0.87 & 0.41 & 0.93 & 0.98 & -0.98\\
+\rowcolor{gray!6}  groceries & -0.48 & 0.87 & 1.00 & 0.44 & 0.87 & 0.87 & -0.87\\
+parks & -0.27 & 0.41 & 0.44 & 1.00 & 0.49 & 0.39 & -0.41\\
+\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.49 & 1.00 & 0.93 & -0.95\\
 \addlinespace
-work & -0.69 & 0.98 & 0.87 & 0.44 & 0.94 & 1.00 & -0.99\\
-\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.86 & -0.45 & -0.95 & -0.99 & 1.00\\
+work & -0.69 & 0.98 & 0.87 & 0.39 & 0.93 & 1.00 & -0.99\\
+\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.87 & -0.41 & -0.95 & -0.99 & 1.00\\
 \bottomrule
@@
 
-A regression model uses the log of new daily cases as the dependent
-variable. The covariates are parks- and residential-related mobility,
-which enter the regression in the form of a second order polynomial
-expansion. In addition, the date is introduced to account for the
-temporal trend of the pandemic. Finally, an indicator variable for the
-state of New York and an interaction with the date are used to
-distinguish the unusually high incidence of the disease there. The
-results appear in Table \ref{tab:model-results}. The model provides a
-good fit to the data and all variables reported are significant at
-\(p<0.05\) or better.
+A regression model is estimated with the log of new daily cases as the
+dependent variable. The covariates enter the regression in the form of a
+second order polynomial expansion. In addition, the date (centered on
+April 5) is introduced to account for the temporal trend of the
+pandemic. Finally, an indicator variable for the state of New York and
+an interaction with the date are used to distinguish the unusually high
+incidence of the disease there. The results appear in Table
+\ref{tab:model-results}. The model provides a good fit to the data and
+all variables reported are significant at \(p<0.05\) or better.
 
 There is an overall temporal trend that indicates a growing number of
-cases over time, but at a decreasing rate (see negative sign of
-date\^{}2). New York has on average more new daily cases than the rest
-of the states, but this has also declined over time (see negative sign
-of NY x date). Visualization is the most effective way to understand the
-trend according to the mobility indicators and date. Figure
-\ref{fig:prediction-plots} shows the prediction surfaces on three
-different dates: March 21, when the first states began implementing
-stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
-more days later, and at a time when some states are considering relaxing
-stay-at-home orders. On March 21 there were still only minor departures
-from baseline mobility (recall that these are temporally lagged); the
-prediction surface at this point is relatively flat. This changes by
-April 5, when large increases in residential-based mobility are seen,
-along with greater variations in parks-based mobility. The prediction
-surface indicates an expectation of a greater number of new daily cases
-as both classes of mobility increase. On the last date examined, April
-20, the trend becomes more steep for park-based mobility, even as this
-indicator continues to display large variations from the baseline in
-both directions. In general, higher levels of mobility tend to be
-associates with a higher number of new daily cases.
+cases over time, at an accelerating rate (see positive sign of
+date\^{}2). Mobility related to parks and to work both tend to increase
+the number of new cases; the negative sign for their interaction is
+indicative of the trade-offs between these two forms of mobility and
+their impact on new cases. The influence of parks-related mobility was
+relatively weaker early in the pandemic (negative sign of the parks x
+date term) but has become more important over time (positive sign of the
+parks x date\^{}2 term). The opposite happens with work-related
+mobility, which started with a greater effect (positive sign of work x
+date term), but whose impact has declined over time (negative sign of
+work x date\^{}2 term) New York has on average more new daily cases than
+the rest of the states, but this has declined over time (see negative
+sign of NY x date).
 
-\begin{table}[!h]
+\begin{table}[H]
 
@@
 \centering
+\fontsize{7}{9}\selectfont
 \begin{tabular}[t]{lcc}
@@
 \midrule
-\rowcolor{gray!6}  date & 7.9619 & <0.0001\\
-date\textasciicircum{}2 & -0.0411 & <0.0001\\
-\rowcolor{gray!6}  parks\textasciicircum{}2 & 1.4017 & 0.0251\\
-parks & -37.2492 & <0.0001\\
-\rowcolor{gray!6}  parks x residential & 39.0902 & <0.0001\\
+\rowcolor{gray!6}  date & 0.1085 & <0.001\\
+date\textasciicircum{}2 & 0.0070 & <0.001\\
+\rowcolor{gray!6}  parks\textasciicircum{}2 & 2.1756 & <0.001\\
+parks & 8.9203 & <0.001\\
+\rowcolor{gray!6}  parks x work & -23.6010 & <0.001\\
 \addlinespace
-residential & -286.7568 & <0.0001\\
-\rowcolor{gray!6}  residential\textasciicircum{}2 & 249.9901 & <0.0001\\
-parks x date & -0.1412 & <0.0001\\
-\rowcolor{gray!6}  residential x date & -6.9180 & <0.0001\\
-residential x date\textasciicircum{}2 & 0.0365 & <0.0001\\
+work & 6.3156 & 0.0035\\
+\rowcolor{gray!6}  work\textasciicircum{}2 & 12.0539 & <0.001\\
+parks x date & -0.2133 & <0.001\\
+\rowcolor{gray!6}  parks x date\textasciicircum{}2 & 0.0079 & <0.001\\
+work x date & 0.1376 & 0.0272\\
 \addlinespace
-\rowcolor{gray!6}  NY & 6.9864 & <0.0001\\
-NY x date & -0.0557 & 0.0022\\
+\rowcolor{gray!6}  work x date\textasciicircum{}2 & -0.0238 & <0.001\\
+NY & 3.5466 & <0.001\\
+\rowcolor{gray!6}  NY x date & -0.0527 & <0.001\\
 \bottomrule
 \multicolumn{3}{l}{\textit{Note: }}\\
-\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.821}\\
-\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.82}\\
-\multicolumn{3}{l}{Standard Error $\sigma$= 2.102}\\
+\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.831}\\
+\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.83}\\
+\multicolumn{3}{l}{Standard Error $\sigma$= 2.081}\\
 \end{tabular}
 \end{table}
+
+Visualization is the most effective way to understand the trend
+according to the mobility indicators and date. Figure
+\ref{fig:prediction-plots} shows the prediction surfaces on three
+different dates: March 21, when the first states began implementing
+stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
+more days later, and at a time when some states started to consider
+relaxing stay-at-home orders.
+
+On March 21 there were still only minor departures from baseline
+mobility (recall that these are temporally lagged); the prediction
+surface at this point is relatively flat. This changes by April 5, when
+work-based mobility has declined substantially. Although every state
+registers lower work-based mobility, there are large variations in
+parks-based mobility, with some states seeing increases of up to 60\%
+for this class of mobility.
+
+The prediction surface indicates an expectation of a greater number of
+new daily cases as either class of mobility increases, but the effect is
+not linear. On the last date examined, April 20, the trend becomes more
+steep for park-based mobility, even as this indicator continues to
+display large variations from the baseline in both directions. The white
+dashed lines in the plots are the folds of the saddles, and represent,
+for each date, the combination of parks- and work-related mobility
+levels that tended to minimize the emergence of new cases.
 
@@
 \caption{\label{fig:prediction-plots}Prediction surfaces at three points
-during the pandemic according to the model; the dots are scatterplots of
-the park- and residential-mobility indicators of the states on that
-date.}
+during the pandemic according to the model; the dots are a scatterplot
+of the parks- and work-related mobility indicators of the states on that
+date; the white dashed line is the fold of the saddle.}
 \end{figure}
 
-The results suggest the potential of GCMR to investigate the spread of
+The results suggest that over time the benefits of reduced work-related
+mobility can be offset by parks-related mobility. For example, on March
+21, New Jersey and Idaho had similar levels of park-related mobility. By
+April 20, New Jersey had substantially reduced park-related mobility,
+whereas Idaho's was even higher than in March. The estimated and actual
+number of new cases grew in the intervening period; however, New
+Jersey's growth in new cases (from March 21 to April 20) was only 894\%,
+whereas Idaho's was 1030\%.
+
+These results suggest the potential of GCMR to investigate the spread of
 COVID-19, but also point at some limitations. The baseline level is not
-defined in a metric that is amenable to policy development (Google
-defines ``mobility'' as an aggregated score of visits and length of stay
-at places.) For example, it is not clear precisely what is residential
-mobility: is it visits with friends and relatives, or mobility in the
-vicinity of the place of residency? Without a clearer understanding of
-these variables, their use can suggest trends, but their potential for
-applied policy analysis appears to be more limited.
+defined in a metric that is amenable to policy development (e.g.,
+person-km travelled). Without a clearer understanding of the absolute
+levels of these variables, their use can suggest trends, but their
+potential for applied policy analysis appears to be more limited.
 

### 2020-05-11 README.Rmd
Updated README.Rmd with author information
Second version of paper after first round of anonymous reviews.

@@
 
+Antonio Paez (paezha@mcmaster.ca)  
+School of Geography and Earth Science  
+McMaster University  
+1280 Main Street West, Hamilton, ON L8S 4K1  
+Canada
+
+(revised versin of paper after anonymous reviews submitted to Transport Findings on May 11, 2020)
+
 ## Abstract
@@
 
-The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders.
-Concurrently, numerous efforts exist to track the progress and the impact of the pandemic, and new sources of data include the recently-released Google Community Mobility Reports [GCMR](https://www.google.com/covid19/mobility/), as well as The New York Times repository of [COVID-19 data](https://github.com/nytimes/covid-19-data).
+The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance.
+Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic.
+As a result, there are new sources of data such as the recently-released Google Community Mobility Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as The New York Times repository of COVID-19 data\footnote{https://github.com/nytimes/covid-19-data}.
 These two open data sets offer novel opportunities to investigate in quasi-real time the relationship between mobility patterns and transmission of COVID-19.
@@
 
-The source document for this paper is an `R` markdown document available from the following repository:
-
-https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/tree/fcd055878020bf567cd4deabb16c8056189eb45c/Covid-19-Google-CMR-US
+This paper is a reproducible research document.
+The source is an `R` markdown file available in a public repository\footnote{See folder Covid-19-Google-CMR-US in \url{https://github.com/paezha/Google-Mobility-Reports-and-COVID-19-US/}}.
 
@@
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
-  kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
+  column_spec(2, width = "12em") %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
         caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
-  #kable_styling(latex_options = c("striped", "scale_down", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
            pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
-  mutate(pval = ifelse(pval >= 0.0001, round(pval, 4), "<0.0001")) %>%
+  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
   kable(#"latex",
@@
         col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
-  #kable_styling(latex_options = c("striped", "hold_position")) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position")) %>%
   footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),

### 2020-05-19 Covid-19-Google-CMR-US/Response-to-Reviewers-v3.Rmd
v3 Revisions after second round of anonymous reviews.

---
title: "Response to Reviewers"
author: "Antonio Paez"
date: "5/11/2020"
output: pdf_document
---

```{r load-packages, include=FALSE}
#Load the packages needed to read and work with the data
library(adespatial)
library(countrycode)
library(gridExtra)
library(interactions)
library(jtools)
library(kableExtra)
library(lubridate)
library(metR)
library(plotly)
library(RCurl)
library(sf)
library(spdep)
library(tidyverse)
library(tigris)
library(usmap)
library(wppExplorer)
```

```{r load-saved-gcmr-data, include=FALSE}
#Load data:
load("gmobility.RData")
```

```{r extract-gcmr-for-us, include=FALSE}
gmobility <- gmobility %>%
  filter(country_region == "United States") %>%
  droplevels()
```

```{r prepare-mobility-report-data, include=FALSE}
#`date` is stored as a factor which is not ideal work with times; Change format:
gmobility <- gmobility %>%
  mutate(date = lubridate::ymd(date)) %>%
  #rename the annoyingly long names of the columns to something more practical
  transmute(date,
            state = as.character(sub_region_1),
         county = as.character(sub_region_2),
         r_and_r = retail_and_recreation_percent_change_from_baseline, 
         g_and_p = grocery_and_pharmacy_percent_change_from_baseline, 
         parks = parks_percent_change_from_baseline,
         transit = transit_stations_percent_change_from_baseline, 
         work = workplaces_percent_change_from_baseline,
         residential = residential_percent_change_from_baseline) %>%
  #convert the mobility indicators to differences from the baseline, so "1" is the baseline value
  mutate(r_and_r = 1 + r_and_r/100,
         g_and_p = 1 + g_and_p/100,
         parks = 1 + parks/100,
         transit = 1 + transit/100,
         work = 1 + work/100,
         residential = 1 + residential/100)
```

```{r filter-state-gcmr-data, include=FALSE}
#To extract the state-level mobility indicators select those records that do not include the label for the county (i.e., sub_region_2 is empty), and that are not national level statistics (i.e., sub_region_1 is not empty)
cmr <- gmobility %>%
  filter(state !="", 
         county == "")
```

```{r load-saved-covid-19-data, include=FALSE}
#Load data:
load("covid19_nyt.RData")
```

```{r coterminous-states, include=FALSE}
#filter states to keep only the coterminous states
c19 <- c19 %>%
  filter(state != "Hawaii",
         state != "Alaska",
         state != "United States Virgin Islands",
         state != "Commonwealth of the Northern Mariana Islands",
         state != "Guam",
         state != "American Samoa",
         state != "Puerto Rico")
```

```{r add-day-variable, include=FALSE}
#add day variable:
c19 <- c19 %>%
  mutate(day = yday(date) - yday("2020-04-05"))
```

```{r calculate-new-daily-cases, include=FALSE}
#To calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total. Also calculate the daily percent change in new cases (pct_change), and the percent increase in total cases (pct_increase):
c19 <- c19 %>%
  group_by(state) %>%
  mutate(new_cases = cases - lag(cases, 1),
         pct_change = (new_cases - lag(new_cases, 1))/lag(new_cases, 1),
         pct_increase = (cases - lag(cases, 1))/lag(cases, 1)) %>%
  ungroup() %>%
  #there are two record with negative numbers of new cases, likely corrections (Georgia in 2020-04-12, Idaho in 2020-05-03). Remove record:
  filter(new_cases >= 0)
```

```{r state-population-data, include=FALSE}
#Retrieve population data:
population <- statepop %>%
  transmute(pop_2015, state = full)
```

```{r join-population-to-covid-19-data, include=FALSE}
#Join population to covid-19 data:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(population, by = c("state" = "state"))
```

```{r calculate-incidence, include=FALSE}
#Calculate the incidence of reported cases by 100,000 population:
c19 <- c19 %>%
  mutate(incidence = cases/(pop_2015/100000),
         log_incidence = log(incidence))
```

```{r calculate-log-new-cases, include=FALSE}
#calculate log of new cases
c19 <- c19 %>% 
  mutate(log_new_cases = log(new_cases + 0.001))
```

```{r join-gcmr-and-covid-19-data, include=FALSE}
#Join the tables:
c19 <- c19 %>% 
  mutate(state = as.character(state)) %>%
  left_join(cmr, by = c("state" = "state", "date" = "date"))
```

```{r calculate-lagged-mobility-indicators, include=FALSE}
#The potential salutary effect of the lockdown (measured in terms of reduced mobility) are not immediate, given the incubation period of the disease. The incubation period has been estimated to be between 2 and 11 days (95% interval) by [Lauer et al. (2020)](https://annals.org/aim/fullarticle/2762808/incubation-period-coronavirus-disease-2019-covid-19-from-publicly-reported). If we wanted to explore the way reduced mobility correlates with number of new cases, we would need to lag the mobility variables by some period.

# Lagged retail and recreation: 11 day moving average
r_and_r_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$r_and_r, 12) + 
                      lag(.x$r_and_r, 11) + 
                      lag(.x$r_and_r, 10) + 
                      lag(.x$r_and_r, 9) + 
                      lag(.x$r_and_r, 8) + 
                      lag(.x$r_and_r, 7) + 
                      lag(.x$r_and_r, 6) + 
                      lag(.x$r_and_r, 5) + 
                      lag(.x$r_and_r, 4) + 
                      lag(.x$r_and_r, 3) + 
                      lag(.x$r_and_r, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_r_and_r_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_r_and_r_lag11)

# Lagged groceries and pharmacies: 11 day moving average
g_and_p_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$g_and_p, 12) + 
                      lag(.x$g_and_p, 11) + 
                      lag(.x$g_and_p, 10) + 
                      lag(.x$g_and_p, 9) + 
                      lag(.x$g_and_p, 8) + 
                      lag(.x$g_and_p, 7) + 
                      lag(.x$g_and_p, 6) + 
                      lag(.x$g_and_p, 5) + 
                      lag(.x$g_and_p, 4) + 
                      lag(.x$g_and_p, 3) + 
                      lag(.x$g_and_p, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_g_and_p_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_g_and_p_lag11)

# Lagged parks: 11 day moving average
parks_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$parks, 12) + 
                      lag(.x$parks, 11) + 
                      lag(.x$parks, 10) + 
                      lag(.x$parks, 9) + 
                      lag(.x$parks, 8) + 
                      lag(.x$parks, 7) + 
                      lag(.x$parks, 6) + 
                      lag(.x$parks, 5) + 
                      lag(.x$parks, 4) + 
                      lag(.x$parks, 3) + 
                      lag(.x$parks, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_parks_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_parks_lag11)

# Lagged groceries and pharmacies: 11 day moving average
transit_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$transit, 12) + 
                      lag(.x$transit, 11) + 
                      lag(.x$transit, 10) + 
                      lag(.x$transit, 9) + 
                      lag(.x$transit, 8) + 
                      lag(.x$transit, 7) + 
                      lag(.x$transit, 6) + 
                      lag(.x$transit, 5) + 
                      lag(.x$transit, 4) + 
                      lag(.x$transit, 3) + 
                      lag(.x$transit, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_transit_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_transit_lag11)

# Lagged groceries and pharmacies: 11 day moving average
work_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$work, 12) + 
                      lag(.x$work, 11) + 
                      lag(.x$work, 10) + 
                      lag(.x$work, 9) + 
                      lag(.x$work, 8) + 
                      lag(.x$work, 7) + 
                      lag(.x$work, 6) + 
                      lag(.x$work, 5) + 
                      lag(.x$work, 4) + 
                      lag(.x$work, 3) + 
                      lag(.x$work, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_work_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_work_lag11)

# Lagged residential: 11 day moving average
residential_lag11 <- cmr %>% 
  group_by(state) %>%
  group_modify(~ {((lag(.x$residential, 12) + 
                      lag(.x$residential, 11) + 
                      lag(.x$residential, 10) + 
                      lag(.x$residential, 9) + 
                      lag(.x$residential, 8) + 
                      lag(.x$residential, 7) + 
                      lag(.x$residential, 6) + 
                      lag(.x$residential, 5) + 
                      lag(.x$residential, 4) + 
                      lag(.x$residential, 3) + 
                      lag(.x$residential, 2))/11) %>%
      enframe()}) %>%
  ungroup() %>%
  transmute(Mean_residential_lag11 = value)  %>%
  cbind(cmr) %>%
  transmute(state = as.character(state), 
            date, Mean_residential_lag11)
```

```{r join-lagged-indicators, include=FALSE}
#join lagged mobility indicators to table:
c19 <- c19 %>%
  mutate(state = as.character(state)) %>%
  left_join(r_and_r_lag11, by = c("state" = "state", "date" = "date")) %>%
  left_join(g_and_p_lag11, by = c("state" = "state", "date" = "date"))  %>%
  left_join(parks_lag11, by = c("state" = "state", "date")) %>%
  left_join(transit_lag11, by = c("state" = "state", "date")) %>%
  left_join(work_lag11, by = c("state" = "state", "date")) %>%
  left_join(residential_lag11, by = c("state" = "state", "date"))
```

```{r remove-nas, include=FALSE}
#Remove all the "NAs" to obtain a clean table with COVID-19 and mobility information:
c19 <- c19 %>%
  drop_na()
```

```{r prepare-variables-for-analysis, include=FALSE}
#replace mobility indicators by their lagged moving averages
c19 <- c19 %>% 
  mutate(retail = Mean_r_and_r_lag11,
         groceries = Mean_g_and_p_lag11,
         parks = Mean_parks_lag11,
         parks2 = Mean_parks_lag11^2,
         transit = Mean_transit_lag11,
         work = Mean_work_lag11,
         work2 = work^2,
         residential = Mean_residential_lag11,
         residential2 = Mean_residential_lag11^2,
         parksXresidential = parks * residential,
         parksXwork = parks * work,
         day2 = day^2,
         parksXday = parks * day,
         parksXday2 = parks * day2,
         workXday = work * day,
         workXday2 = work * day2,
         residentialXday = residential * day,
         residentialXday2 = residential * day2,
         NY = ifelse(state == "New York", 1, 0),
         NYxday = NY * day)
```

# Reviewer 1

Thanks for the revisions. I think they make the paper stronger.

>\textcolor{blue}{Thank you for your positive assessment of the paper and comments.}

I have a question on the rationale for using daily new cases (log transformed) as a dependent variable. Given the large differences in population size, urban areas and densities across states, wouldn’t using percentage change in new cases day to day be a more appropriate dependent variable (or some moving average of it)? Such a setup removes the population size effect...I would suggest the authors at least test the percent variable and see if it allows for a simpler specification. Or alternately, explain why log(new cases) is a valid measure to answer their question given differences across states that are not accounted for.

>\textcolor{blue}{Thank you for this comment. I understand percentage change in new cases for state $i$ and day $t$ to mean the following:}

$$
pct\_change_{i,t} = \frac{new\_cases_{i,t} - new\_cases_{i,t-1}}{new\_cases_{i,t-1}}
$$

>\textcolor{blue}{Alas, as seen in Table \ref{tab:correlation-analysis-pct-change} below, percent change correlates poorly with the independent variables, and the low correlations translate into a poor fit with non-significant coefficients (see Table \ref{tab:model-results-pct-change}). This is because the daily percent change in new cases is very noisy and not a good indication of a trend: even small changes (say from +1 to -1 case change from day to day) change the sign of the variable.}

```{r check-correlations-pct-change, echo=FALSE}
#Check the correlation between the mobility indicators and day-to-day percent change in new cases:
c19 %>% 
  filter(is.finite(pct_change)) %>%
  select(pct_change, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-pct-change}Simple correlation between day to day percent change in new Cases and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r model-estimation-pct-change, include=FALSE}
mod_pct_change <- lm(pct_change ~  0 +
                         day +
                         day2 +
                         parks2 + 
                         parks + 
                         parksXwork +
                         work +
                         work2 +
                         parksXday + 
                         parksXday2 +
                         workXday +
                         workXday2 +
                         NY + 
                         NYxday, 
                       data = c19 %>% filter(is.finite(pct_change))
)
summ(mod_pct_change)
```

```{r model-results-pct-change, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x day^2",
                        "work x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod_pct_change)),
           pval = as.numeric(summary(mod_pct_change)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results-pct-change}Dependent variable is day-to-day percent change in new cases.",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod_pct_change)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod_pct_change)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod_pct_change)$sigma, 3))),
           escape = FALSE)
```

>\textcolor{blue}{Instead of day-to-day percent change in new cases, we could define a variable that measures the percentage increase in total cases:}

$$
pct\_increase_{i,t} = \frac{cases_{i,t} - cases_{i,t-1}}{cases_{i,t-1}}
$$

>\textcolor{blue}{Again, a variable like this would normalize for the caseload in each state. The correlations for percent increase are shown in Table \ref{tab:correlation-analysis-pct-increase}, were we see that they are considerably higher than for day-to-day percent change in cases, but still lower than for log of new cases. If we use daily percent increase in total cases as the dependent variable, the resulting model is somewhat more parsimonious but the goodness of fit is also quite a bit lower than the model with log of new cases (see Table \ref{tab:model-results-pct-increase}).}

```{r check-correlations-pct-increase, echo=FALSE}
#Check the correlation between the mobility indicators and daily percent increase in total cases:
c19 %>% 
  select(pct_increase, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-pct-increase}Simple correlation between daily percent increase in total cases and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r model-estimation-pct-increase, include=FALSE}
mod_pct_increase <- lm(pct_increase ~  0 +
                         #day +
                         #day2 +
                         #parks2 + 
                         parks + 
                         parksXwork +
                         work +
                         work2 +
                         parksXday + 
                         parksXday2 +
                         #workXday +
                         workXday2 +
                         NY + 
                         NYxday, 
                       data = c19)
summ(mod_pct_increase)
```

```{r model-results-pct-increase, echo = FALSE}
data.frame(Variable = c(#"date",
                        #"date^2",
                        #"parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x date^2",
                        "NY",
                        "NY x date"),
           Coefficient = as.numeric(coefficients(mod_pct_increase)),
           pval = as.numeric(summary(mod_pct_increase)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results-pct-increase}Dependent variable is daily percent increase in total cases.",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod_pct_increase)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod_pct_increase)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod_pct_increase)$sigma, 3))),
           escape = FALSE)
```

>\textcolor{blue}{The examples above show that using the day-to-day percent daily change in new cases and percent daily increase in total cases does not result in better models. The point about controlling for population still stands. This made me think of a different dependent variable, namely incidence per 100,000:}

$$
incidence_{i,t} = \frac{Cases_{i,t}}{population_{i}/100,000}
$$

>\textcolor{blue}{Incidence normalizes the number of cases and accounts for the size of the population at risk. This variable is log-transformed to scale it (it has a long tail, as you would expect). The correlations with the mobility indicators are shown in Table \ref{tab:correlation-analysis-log-incidence}. Using the log of incidence as the dependent variable results in a somewhat more parsimonious model with better goodness of fit (see Table \ref{tab:model-results-log-incidence}). Based on these results, this is the model that I report in the paper now. It is worthwhile noting that qualitatively the results are very similar, but with a slightly smaller model that also performs better.}

```{r check-correlations-log-incidence, echo=FALSE}
#Check the correlation between the mobility indicators and log of incidence:
c19 %>% 
  select(log_incidence, retail, groceries, parks, transit, work, residential) %>%
  cor(use = "pairwise.complete.obs") %>%
  kable("latex",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:correlation-analysis-log-incidence}Simple correlation between log(incidence) and the mobility indicators")  %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = "All mobility indicators are lagged 11-day moving averages")
```

```{r model-estimation-log-incidence, include=FALSE}
mod1 <- lm(log_incidence ~  0 +
                         day +
                         day2 +
                         parks2 + 
                         parks + 
                         parksXwork +
                         work +
                         work2 +
                         parksXday + 
                         parksXday2 +
                         #workXday +
                         workXday2 +
                         NY, #+ 
                         #NYxday, 
                       data = c19)
summ(mod1)
```

```{r model-results-log-incidence, echo = FALSE}
data.frame(Variable = c("date",
                        "date^2",
                        "parks^2",
                        "parks",
                        "parks x work",
                        "work",
                        "work^2",
                        "parks x date",
                        "parks x date^2",
                        "work x date^2",
                        "NY"),
           Coefficient = as.numeric(coefficients(mod1)),
           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
  kable("latex",
        booktabs = TRUE,
        digits = 4,
        caption = "\\label{tab:model-results-log-incidence}Dependent variable is log(Incidence).",
        align = c("l", "c", "c"),
        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
  kable_styling(latex_options = c("striped"), font_size = 7) %>%
  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
           escape = FALSE)
```

I am also concerned that almost all the variables in the model have a quadratic form and by the number of interactions in the model. Is this an artifact of not controlling for the population size effect?

>\textcolor{blue}{The quadratic form is to capture trends in attribute space. As the performance of the model shows, these terms are effective at this task. And seeing how the terms are significant even after controlling for population size (i.e., modelling the incidence per 100,000 instead of new cases), I hope that your concerns will be allayed.}

>\textcolor{blue}{Finally, I have taken the opportunity of this revision to update the analysis to May 5, 2020.}

### 2020-05-19 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.Rmd
v3 Revisions after second round of anonymous reviews.

@@
 ---
-title: Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
+title: Using Google Community Mobility Reports to investigate the incidence of COVID-19 in the United States
 author:
@@
 These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline.
-This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19.
-Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US.
+This novel source of data offers an opportunity to investigate potential correlations between mobility and incidence of COVID-19.
+Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and incidence of COVID-19 by state in the US.
 The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.
 
-journal: "Transportation Findings"
+journal: "Transport Findings"
 date: "`r Sys.Date()`"
@@
 ```{r calculate-new-daily-cases, include=FALSE}
-#o calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total:
+#To calculate the number of new daily cases by state, we need to lag the Cases variable and substract the previous total.
+Also calculate the daily percent change in new cases (pct_change), and the percent increase in total cases (pct_increase):
 c19 <- c19 %>%
   group_by(state) %>%
-  mutate(new_cases = cases - lag(cases, 1)) %>%
+  mutate(new_cases = cases - lag(cases, 1),
+         pct_change = (new_cases - lag(new_cases, 1))/lag(new_cases, 1),
+         pct_increase = (cases - lag(cases, 1))/lag(cases, 1)) %>%
   ungroup() %>%
-  #there is one record with a negative number of new cases, likely a correction (Georgia in 2020-04-12). Remove record:
+  #there are two record with negative numbers of new cases, likely corrections (Georgia in 2020-04-12, Idaho in 2020-05-03). Remove record:
   filter(new_cases >= 0)
@@
 
-The main policy tool to control the spread of the COVID-19 pandemic has been stay-at-home orders, which in the United States have been implemented on a state-by-state basis, with considerable variations in compliance.
+The main policy tool to control the spread of the COVID-19 pandemic has been restrictions to non-essential travel in the form of stay-at-home orders.
+In the United States, such orders have been implemented on a state-by-state basis with considerable variations in compliance.
 Concurrently, numerous initiatives have been developed to track the progress and the impact of the pandemic.
@@
 
-This paper investigates the potential of Google Community Mobility Reports to asses the impact of mobility on COVID-19. The following questions are posed:
+This paper investigates the potential of Google Community Mobility Reports to asses the impact of mobility on the incidence of COVID-19. The following questions are posed:
 
-- Do changes in mobility according to GCMR correlate with the transmission of COVID-19?
+- Do changes in mobility according to GCMR correlate with the incidence of COVID-19?
 - And if so, what do we learn about mobility and the spread of the disease?
@@
 GCMR use aggregated and anonymized data to chart changes in mobility with respect to different classes of places (see Table \ref{tab:descriptive-statistics}).
-Mobility indicators are calculated based on the frequency and length of visits.
+Mobility indicators are calculated based on the frequency and length of visits to places.
 The reports give percentage change from a baseline level, which corresponds to the median value of mobility of identical days of the week during the period between January 3 and Feb 6, 2020.
@@
 ```{r descriptive-statistics, echo=FALSE}
-data.frame(Variable = c("New Cases",
+data.frame(Variable = c("Incidence",
                         "date",
@@
                         "residential"),
-           Definition = c("New Daily Cases of COVID-19",
+           Definition = c("Total cases of COVID-19 divided by population (in 100,000s)",
                           "Date",
@@
                           "Mobility trends for places of residence"),
-           min = c(as.character(min(c19$new_cases)),
+           min = c(as.character(round(min(c19$incidence), 2)),
                    as.character(min(c19$date)),
@@
                    as.character(round(min(c19$residential), 2))),
-           median = c(as.character(median(c19$new_cases)),
+           median = c(as.character(round(median(c19$incidence), 2)),
                       as.character(median(c19$date)),
@@
                       as.character(round(median(c19$residential), 2))),
-           max = c(as.character(max(c19$new_cases)),
+           max = c(as.character(round(max(c19$incidence), 2)),
                    as.character(max(c19$date)),
@@
                    as.character(round(max(c19$residential), 2))),
-           sd = c(as.character(round(sd(c19$new_cases), 2)),
+           sd = c(as.character(round(sd(c19$incidence), 2)),
                   as.character(""),
@@
 
-For analysis, all mobility indicators are centered so that the value of 1 is the base mobility, and a 0.01 deviation corresponds to a 1% change.
+For analysis, all mobility indicators are centered so that the value of 1 is the baseline mobility, and a 0.01 deviation corresponds to a 1% change.
 The incubation time of the disease is between 2 and 12 days [95% interval; see @Lauer2020incubation].
-Given this, it is to be expected that any changes in mobility will have a lagged effect (if any) on the discovery of new cases.
+Given this, it is to be expected that any changes in mobility will have a lagged effect on the discovery of new cases.
 For this reason, lagged moving averages of the mobility indicators are calculated.
@@
 The lagged indicators are calculated as the mean of the mobility indicator using the values from date-minus-12-days to date-minus-2-days.
-Furthermore, using the cumulative number of reported COVID-19 cases, the total number of new daily cases is calculated.
-This variable (log-transformed after adding a small constant) is paired with the corresponding lagged moving average of the mobility indicators.
-The log-transformation is useful to avoid negative values of daily new cases when making predictions.
+Furthermore, using the cumulative number of reported COVID-19 cases, the incidence is calculated after dividing by the population of the state (in 100,000s).
+This variable (log-transformed) is paired with the corresponding lagged moving average of the mobility indicators.
+The log-transformation is useful to avoid negative values of incidence when making predictions.
 Table \ref{tab:descriptive-statistics} shows the descriptive statistics of the data set.
@@
 
-Table \ref{tab:correlation-analysis} shows that the mobility indicators are highly correlated with each other.
+Table \ref{tab:correlation-analysis-log-incidence} shows that the mobility indicators are highly correlated with each other.
 Two variables are selected for multivariate analysis: parks- and work-related mobility.
-Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables.
+Work has a high correlation with the outcome variable, and its correlation with parks is relatively weak, which increases the information content of the two variables in multivariate analysis.
 Furthermore, parks- and work-related mobility represent two dimensions of out-of-home activities: mandatory and discretionary travel.
 
-```{r check-correlations, echo=FALSE}
+```{r check-correlations-log-new-cases, eval=FALSE, include=FALSE}
 #Check the correlation between the mobility indicators and incidence:
@@
 
-A regression model is estimated with the log of new daily cases as the dependent variable.
+```{r check-correlations-log-incidence, echo=FALSE}
+#Check the correlation between the mobility indicators and log of incidence:
+c19 %>% 
+  select(log_incidence, retail, groceries, parks, transit, work, residential) %>%
+  cor(use = "pairwise.complete.obs") %>%
+  kable("latex",
+        booktabs = TRUE,
+        digits = 2,
+        caption = "\\label{tab:correlation-analysis-log-incidence}Simple correlation between log(incidence) and the mobility indicators")  %>%
+  kable_styling(latex_options = c("striped", "scale_down"), font_size = 7) %>%
+  footnote(general = "All mobility indicators are lagged 11-day moving averages")
+```
+
+A regression model is estimated with the log of incidence as the dependent variable.
 The covariates enter the regression in the form of a second order polynomial expansion.
 In addition, the date (centered on April 5) is introduced to account for the temporal trend of the pandemic.
-Finally, an indicator variable for the state of New York and an interaction with the date are used to distinguish the unusually high incidence of the disease there.
-The results appear in Table \ref{tab:model-results}.
-The model provides a good fit to the data and all variables reported are significant at $p<0.05$ or better.
+Finally, an indicator variable for the state of New York is used to distinguish the unusually high incidence of the disease there.
+The results of the model are shown in Table \ref{tab:model-results-log-incidence}.
+The model provides a good fit to the data and all variables reported are significant at $p<0.10$ or better.
 
-There is an overall temporal trend that indicates a growing number of cases over time, at an accelerating rate (see positive sign of date^2).
-Mobility related to parks and to work both tend to increase the number of new cases; the negative sign for their interaction is indicative of the trade-offs between these two forms of mobility and their impact on new cases.
-The influence of parks-related mobility was relatively weaker early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term).
-The opposite happens with work-related mobility, which started with a greater effect (positive sign of work x date term), but whose impact has declined over time (negative sign of work x date^2 term) New York has on average more new daily cases than the rest of the states, but this has declined over time (see negative sign of NY x date).
+There is an overall temporal trend that indicates a growing incidence over time, but at a decelerating rate (see negative sign of date^2).
+Mobility related to parks and to work are both associated with higher incidence of COVID-19, however, the effect of parks-related mobility grows non-linearly (see positive sign of quadratic term), whereas the effect of work-related mobility grows at a decreasing rate (see negative sign of quadratic term).
+Furthermore, the negative sign for the interaction of these two mobility indicators captures the trade-offs between these two forms of mobility and their impact on incidence.
+The influence of parks-related mobility was relatively weak early in the pandemic (negative sign of the parks x date term) but has become more important over time (positive sign of the parks x date^2 term).
+The opposite happens with work-related mobility, the importance of which has declined over time (negative sign of work x date^2 term), but whose impact has declined over time (negative sign of work x date^2 term).
+As seen in the table, incidence of COVID-19 in New York is consistently higher.
 
-```{r model-estimation, include=FALSE}
+```{r model-estimation-log-new-cases, eval = FALSE, include=FALSE}
 #This is the regression model (forcing the regression through the intercept):
@@
 
-```{r model-results, echo = FALSE}
+```{r model-estimation-log-incidence, include=FALSE}
+mod1 <- lm(log_incidence ~  0 +
+             day +
+             day2 +
+             parks2 + 
+             parks + 
+             parksXwork +
+             work +
+             work2 +
+             parksXday + 
+             parksXday2 +
+             #workXday +
+             workXday2 +
+             NY, #+ 
+           #NYxday, 
+           data = c19)
+summ(mod1)
+```
+
+```{r model-results-log-new-cases, eval=FALSE, echo = FALSE}
 data.frame(Variable = c("date",
@@
 
+```{r model-results-log-incidence, echo = FALSE}
+data.frame(Variable = c("date",
+                        "date^2",
+                        "parks^2",
+                        "parks",
+                        "parks x work",
+                        "work",
+                        "work^2",
+                        "parks x date",
+                        "parks x date^2",
+                        "work x date^2",
+                        "NY"),
+           Coefficient = as.numeric(coefficients(mod1)),
+           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
+  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
+  kable("latex",
+        booktabs = TRUE,
+        digits = 4,
+        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(Incidence).",
+        align = c("l", "c", "c"),
+        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
+  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
+                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
+                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
+           escape = FALSE)
+```
+
 Visualization is the most effective way to understand the trend according to the mobility indicators and date.
-Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states started to consider relaxing stay-at-home orders.
+Figure \ref{fig:prediction-plots} shows the prediction surfaces on four different dates at intervals of 15 days: March 21, when the first states began implementing stay-at-home orders; then April 5, two weeks into the lockdown; this is followed by April 20, at a time when some states started to consider relaxing stay-at-home orders; and finally May 5, when some states were reopening and/or letting stay-at-home orders lapse.
 
-On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
+On March 21 there were still only minor departures from the baseline level of mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
 This changes by April 5, when work-based mobility has declined substantially.
 Although every state registers lower work-based mobility, there are large variations in parks-based mobility, with some states seeing increases of up to 60% for this class of mobility.
+By May 5, park-related mobility in some states had increased to 200% of the baseline.
 
-The prediction surface indicates an expectation of a greater number of new daily cases as either class of mobility increases, but the effect is not linear.
-On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
-The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the emergence of new cases.
+The prediction surfaces are hyperbolic paraboloids on any given date, and in general indicate an expectation of higher incidences as either class of mobility increases, but with a progressively steeper trend for park-based mobility over time.
+On the last date examined, May 5, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
+The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the incidence.
 
@@
 c19 %>% 
-  #filter(day == 59) %>% 
-  select(date, day, parks, residential) %>%
+  select(date, day, parks, work) %>%
   summary()
@@
 ```{r create-prediction-grid, include=FALSE}
-new.data = expand.grid(parks = seq(0.36, 1.9, by = 0.01),
+new.data = expand.grid(parks = seq(0.36, 2.28, by = 0.01),
                        work = seq(0.33, 1.08, by = 0.01),
-                       day = seq(-15, 15, 1),
+                       day = seq(-15, 30, 1),
                        NY = c(0, 1)) %>%
@@
 
-new.data <- data.frame(new.data, log_new_cases = predict(mod1, newdata = new.data))
+new.data <- data.frame(new.data, log_incidence = predict(mod1, newdata = new.data))
 ```
@@
 ```{r check-dates-for-examples, include=FALSE}
-c19 %>% filter(date == "2020-04-20" | date == "2020-03-21" | date == "2020-04-05" ) %>% select(date, day) %>% distinct()
+c19 %>% filter(date == "2020-03-21" | 
+                 date == "2020-04-05"| 
+                 date == "2020-04-20" | 
+                 date == "2020-05-05") %>% 
+  select(date, day) %>% distinct()
 ```
@@
 ```{r check-mobility-indicators-on-2020-03-21, include=FALSE}
-c19 %>% filter(date == "2020-03-21") %>% select(day, parks, work) %>% summary()
+#Check the range of the mobility indicators on this date, to use in the plot
+c19 %>% 
+  filter(date == "2020-03-21") %>% 
+  select(day, parks, work) %>% 
+  summary()
 ```
@@
 ```{r plot-03-21, include=FALSE, fig.cap="\\label{fig:plot-03-21}Prediction surface at various levels of the mobility indicators on March 21, 2020"}
-plot_03_21 <- ggplot(data = new.data %>% filter(day == -15, parks >= 0.8527 & parks <= 1.5555, work >= 0.7627 & work <= 0.9409),
+plot_03_21 <- ggplot(data = new.data %>% 
+                       filter(day == -15, 
+                              parks >= 0.8527 & parks <= 1.5555, 
+                              work >= 0.7627 & work <= 0.9409),
                      aes(x = parks,
                          y = work)) +
-  geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
-  geom_abline(intercept = 0.43, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_tile(aes(fill = log_incidence)) + 
+  geom_contour(aes(z = log_incidence), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.75, slope = 0.115, color = "white", linetype = "dashed", size = 0.5) +
+  #geom_abline(intercept = 0.10, slope = 0.64, color = "black", linetype = "dashed", size = 0.5) +
   geom_point(data = c19 %>% filter(day == -15, state != "New York") %>% 
@@
             nudge_y = -0.01) +
-#  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
-  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
+  scale_fill_distiller(name = "log(Incidence)", palette = "Reds", direction = 1) +
   theme_light() +
+  theme(legend.position = "bottom") +
   ggtitle("March 21, 2020")
@@
 ```{r check-mobility-indicators-on-2020-04-05, include=FALSE}
-c19 %>% filter(date == "2020-04-05") %>% select(day, parks, work) %>% summary()
+#Check the range of the mobility indicators on this date, to use in the plot
+c19 %>% 
+  filter(date == "2020-04-05") %>% 
+  select(day, parks, work) %>% 
+  summary()
 ```
@@
                          y = work)) +
-  geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
-  geom_abline(intercept = 0.162, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_tile(aes(fill = log_incidence)) + 
+  geom_contour(aes(z = log_incidence), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.10, slope = 0.64, color = "white", linetype = "dashed", size = 0.5) +
   geom_point(data = c19 %>% filter(day == 0, state != "New York") %>% 
@@
             nudge_y = -0.01) +
-  #scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
-  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
+  scale_fill_distiller(name = "log(Incidence)", palette = "Reds", direction = 1) +
   theme_light() +
+  theme(legend.position = "bottom") +
   ggtitle("April 05, 2020")
@@
 ```{r check-mobility-indicators-on-2020-04-20, include=FALSE}
-c19 %>% filter(date == "2020-04-20") %>% select(day, parks, work) %>% summary()
+#Check the range of the mobility indicators on this date, to use in the plot
+c19 %>% 
+  filter(date == "2020-04-20") %>% 
+  select(day, parks, work) %>% 
+  summary()
 ```
@@
 ```{r plot-04-20, include=FALSE, fig.cap="\\label{fig:plot-04-20}Prediction surface at various levels of the mobility indicators on April 20, 2020"}
-plot_04_20 <- ggplot(data = new.data %>% filter(day == 15, parks >= 0.36 & parks <= 1.6309, work >= 0.3582 & work <= 0.6518),
+plot_04_20 <- ggplot(data = new.data %>%
+                       filter(day == 15, 
+                              parks >= 0.36 & parks <= 1.6309, 
+                              work >= 0.3582 & work <= 0.6518),
                      aes(x = parks,
                          y = work)) +
-  geom_tile(aes(fill = log_new_cases)) + 
-  geom_contour(aes(z = log_new_cases), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
-  geom_abline(intercept = 0.18, slope = 0.45, color = "white", linetype = "dashed", size = 0.5) +
+  geom_tile(aes(fill = log_incidence)) + 
+  geom_contour(aes(z = log_incidence), color = "black", size = 0.1, binwidth = 0.1, alpha = 1) +
+  geom_abline(intercept = 0.10, slope = 0.64, color = "white", linetype = "dashed", size = 0.5) +
   geom_point(data = c19 %>% filter(day == 15, state != "New York") %>% 
@@
             nudge_y = -0.01) +
-#  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1, limits = c(4,15)) +
-  scale_fill_distiller(name = "log(New Cases)", palette = "Reds", direction = 1) +
+  scale_fill_distiller(name = "log(Incidence)", palette = "Reds", direction = 1) +
   theme_light() +
+  theme(legend.position = "bottom") +
   ggtitle("April 20, 2020")
@@
 
-```{r prediction-plots, fig.height=9, echo=FALSE, fig.cap="\\label{fig:prediction-plots}Prediction surfaces at three points during the pandemic according to the model; the dots are a scatterplot of the parks- and work-related mobility indicators of the states on that date; the white dashed line is the fold of the saddle."}
-grid.arrange(plot_03_21, plot_04_05, plot_04_20, nrow = 3)
+```{r check-mobility-indicators-on-2020-05-05, include=FALSE}
+#Check the range of the mobility indicators on this date, to use in the plot
+c19 %>% 
+  filter(date == "2020-05-05") %>% 
+  select(day, parks, work) %>% 
+  summary()
 ```
 
-```{r plot-3d-surface, include=FALSE}
-# Choose day for the plot: April 20 is day == 15, April 5 is day == 0, March 21 is day == -15
+```{r plot-05-05, include=FALSE, fig.cap="\\label{fig:plot-05-05}Prediction surface at various levels of the mobility indicators on May 05, 2020"}
+plot_05_05 <- ggplot(data = new.data %>% 
+                       filter(day == 30, 
+                              parks >= 0.4045 & parks <= 2.2764, 
+                              work >= 0.3955 & work <= 0.7055),
+                     aes(x = parks,
+                         y = work)) +
+  geom_tile(aes(fill = log_incidence)) + 
+  geom_contour(aes(z = log_incidence), color = "black", size = 0.1, binwidth = 0.5, alpha = 1) +
+  geom_abline(intercept = 0.1, slope = 1, color = "white", linetype = "dashed", size = 0.5) +
+  #geom_abline(intercept = 0.10, slope = 0.64, color = "black", linetype = "dashed", size = 0.5) +
+  geom_point(data = c19 %>% filter(day == 30, state != "New York") %>% 
+               select(parks, work)) +
+  geom_text(data = c19 %>% 
+              filter(day == 30,
+                     state == "Arkansas" |
+                       state == "Mississippi" |
+                       state == "Oklahoma" |
+                       state == "Idaho" |
+                       state == "New Jersey" |
+                       state == "Michigan" |
+                       state == "California") %>%
+              select(parks, work, state), 
+            aes(label = state),
+            nudge_y = -0.01) +
+  scale_fill_distiller(name = "log(Incidence)", palette = "Reds", direction = 1) +
+  theme_light() +
+  theme(legend.position = "bottom") +
+  ggtitle("May 05, 2020")
+```
+
+```{r prediction-plots, fig.height=9, echo=FALSE, fig.cap="\\label{fig:prediction-plots}Prediction surfaces at three points during the pandemic according to the model; the dots are a scatterplot of the parks- and work-related mobility indicators of the states on that date; the white dashed line is the fold of the saddle."}
+grid.arrange(plot_03_21, plot_04_05, plot_04_20, plot_05_05, nrow = 2)
+```
+
+```{r plot-3d-surface, eval = FALSE, include=FALSE}
+# Choose day for the plot: March 21 is day == -15, April 5 is day == 0, April 20 is day == 15, May 5 is day 30
 plotly.data <- new.data %>% filter(day == 15) 
@@
 
-The results suggest that over time the benefits of reduced work-related mobility can be offset by parks-related mobility.
-For example, on March 21, New Jersey and Idaho had similar levels of park-related mobility.
-By April 20, New Jersey had substantially reduced park-related mobility, whereas Idaho's was even higher than in March.
-The estimated and actual number of new cases grew in the intervening period; however, New Jersey's growth in new cases (from March 21 to April 20) was only 894%, whereas Idaho's was 1030%.
+```{r calculate-example-change-incidence, include=FALSE}
+example_change_incidence <- c19 %>% 
+  filter(date == "2020-03-21" | date == "2020-04-05", 
+         state == "California" | state == "Idaho") %>%
+  select(date, state, incidence)  %>%
+  pivot_wider(names_from = state, values_from = incidence) %>%
+  mutate(change_ID = (Idaho - lag(Idaho, 1))/lag(Idaho, 1),
+         change_CA = (California - lag(California, 1))/lag(California, 1)) %>%
+  drop_na() %>%
+  select(starts_with("change_"))
+```
 
-These results suggest the potential of GCMR to investigate the spread of COVID-19, but also point at some limitations.
+The results suggest that over time the benefits of reduced work-related mobility can be easily offset by parks-related mobility.
+For example, California has consistently registered lower levels of parks-related mobility whereas Idaho has had high levels of this kind of mobility throughout the pandemic.
+The incidence of COVID-19 grew in the intervening period; however, between March 21 and May 5 growth in incidence in California was `r round(example_change_incidence$change_CA * 100, 2)`% whereas Idaho's growth in incidence over the same period was `r round(example_change_incidence$change_ID * 100, 2)`%.
+
+These results suggest the potential of GCMR to investigate the potential effects of mobility on the incidence of COVID-19.
+In particular, growth appears to be more strongly driven by parks-related mobility.
+In terms of the use of these mobility indicators, there are some limitations that must be acknowledged.
 The baseline level is not defined in a metric that is amenable to policy development (e.g., person-km travelled).
-Without a clearer understanding of the absolute levels of these variables, their use can suggest trends, but their potential for applied policy analysis appears to be more limited.
+Without a clearer understanding of the absolute levels of these variables, these indicators are useful for inference and perhaps short-term forecasting, but their potential for applied policy analysis appears to be more limited.
 

### 2020-05-19 Covid-19-Google-CMR-US/Covid-19-Google-CMR-US.tex
v3 Revisions after second round of anonymous reviews.

@@
 
-  \journal{Transportation Findings} % Sets Journal name
+  \journal{Transport Findings} % Sets Journal name
 
@@
             pdfauthor={},
-            pdftitle={Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States},
+            pdftitle={Using Google Community Mobility Reports to investigate the incidence of COVID-19 in the United States},
             colorlinks=false,
@@
 
-  \title{Using Google Community Mobility Reports to investigate the growth of
+  \title{Using Google Community Mobility Reports to investigate the incidence of
 COVID-19 in the United States}
@@
   source of data offers an opportunity to investigate potential
-  correlations between mobility and transmission of COVID-19. Using data
-  from the New York Times on COVID-19 cases and GCMR, this paper presents
-  an analysis of mobility levels and new daily reported cases of COVID-19
-  by state in the US. The results provide insights about the utility and
-  interpretability of GCMR for COVID-19 research and decision-making.
+  correlations between mobility and incidence of COVID-19. Using data from
+  the New York Times on COVID-19 cases and GCMR, this paper presents an
+  analysis of mobility levels and incidence of COVID-19 by state in the
+  US. The results provide insights about the utility and interpretability
+  of GCMR for COVID-19 research and decision-making.
   \end{abstract}
@@
 The main policy tool to control the spread of the COVID-19 pandemic has
-been stay-at-home orders, which in the United States have been
-implemented on a state-by-state basis, with considerable variations in
-compliance. Concurrently, numerous initiatives have been developed to
-track the progress and the impact of the pandemic. As a result, there
-are new sources of data such as the recently-released Google Community
-Mobility Reports
-(GCMR)\footnote{https://www.google.com/covid19/mobility/}, as well as
-The New York Times repository of COVID-19
+been restrictions to non-essential travel in the form of stay-at-home
+orders. In the United States, such orders have been implemented on a
+state-by-state basis with considerable variations in compliance.
+Concurrently, numerous initiatives have been developed to track the
+progress and the impact of the pandemic. As a result, there are new
+sources of data such as the recently-released Google Community Mobility
+Reports (GCMR)\footnote{https://www.google.com/covid19/mobility/}, as
+well as The New York Times repository of COVID-19
 data\footnote{https://github.com/nytimes/covid-19-data}. These two open
@@
 This paper investigates the potential of Google Community Mobility
-Reports to asses the impact of mobility on COVID-19. The following
-questions are posed:
+Reports to asses the impact of mobility on the incidence of COVID-19.
+The following questions are posed:
 
@@
 \item
-  Do changes in mobility according to GCMR correlate with the
-  transmission of COVID-19?
+  Do changes in mobility according to GCMR correlate with the incidence
+  of COVID-19?
 \item
@@
 \ref{tab:descriptive-statistics}). Mobility indicators are calculated
-based on the frequency and length of visits. The reports give percentage
-change from a baseline level, which corresponds to the median value of
-mobility of identical days of the week during the period between January
-3 and Feb 6, 2020. Covid-19 data is compiled by The New York Times based
-on reports from state and local health agencies.
+based on the frequency and length of visits to places. The reports give
+percentage change from a baseline level, which corresponds to the median
+value of mobility of identical days of the week during the period
+between January 3 and Feb 6, 2020. Covid-19 data is compiled by The New
+York Times based on reports from state and local health agencies.
 
@@
 \midrule
-\rowcolor{gray!6}  New Cases & New Daily Cases of COVID-19 & 0 & 90 & 12312 & 1051.67\\
-date & Date & 2020-02-27 & 2020-04-04 & 2020-05-02 & \\
-\rowcolor{gray!6}  retail & Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters & 0.34 & 0.66 & 1.16 & 0.21\\
-groceries & Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies & 0.66 & 0.94 & 1.26 & 0.13\\
-\rowcolor{gray!6}  parks & Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens & 0.36 & 1.15 & 2.1 & 0.26\\
+\rowcolor{gray!6}  Incidence & Total cases of COVID-19 divided by population (in 100,000s) & 0.01 & 56.87 & 1710.06 & 229.76\\
+date & Date & 2020-02-27 & 2020-04-08 & 2020-05-09 & \\
+\rowcolor{gray!6}  retail & Mobility trends for places like restaurants, cafes, shopping centers, theme parks, museums, libraries, and movie theaters & 0.34 & 0.66 & 1.16 & 0.2\\
+groceries & Mobility trends for places like grocery markets, food warehouses, farmers markets, specialty food shops, drug stores, and pharmacies & 0.66 & 0.93 & 1.26 & 0.12\\
+\rowcolor{gray!6}  parks & Mobility trends for places like local parks, national parks, public beaches, marinas, dog parks, plazas, and public gardens & 0.36 & 1.15 & 2.28 & 0.29\\
 \addlinespace
-transit & Mobility trends for places like public transport hubs such as subway, bus, and train stations & 0.24 & 0.72 & 1.14 & 0.23\\
-\rowcolor{gray!6}  work & Mobility trends for places of work & 0.34 & 0.63 & 1.07 & 0.2\\
-residential & Mobility trends for places of residence & 0.97 & 1.14 & 1.27 & 0.08\\
+transit & Mobility trends for places like public transport hubs such as subway, bus, and train stations & 0.24 & 0.7 & 1.14 & 0.22\\
+\rowcolor{gray!6}  work & Mobility trends for places of work & 0.34 & 0.62 & 1.05 & 0.18\\
+residential & Mobility trends for places of residence & 0.98 & 1.14 & 1.27 & 0.07\\
 \bottomrule
@@
 For analysis, all mobility indicators are centered so that the value of
-1 is the base mobility, and a 0.01 deviation corresponds to a 1\%
+1 is the baseline mobility, and a 0.01 deviation corresponds to a 1\%
 change. The incubation time of the disease is between 2 and 12 days
 (95\% interval; see Lauer et al., 2020). Given this, it is to be
-expected that any changes in mobility will have a lagged effect (if any)
-on the discovery of new cases. For this reason, lagged moving averages
-of the mobility indicators are calculated. Furthermore, it is possible
-that mobility and reports of new cases of COVID-19 are endogenous, if
-the public adjust their mobility according to reports of the incidence.
+expected that any changes in mobility will have a lagged effect on the
+discovery of new cases. For this reason, lagged moving averages of the
+mobility indicators are calculated. Furthermore, it is possible that
+mobility and reports of new cases of COVID-19 are endogenous, if the
+public adjust their mobility according to reports of the incidence.
 Therefore, in addition to being consistent with an incubation period,
@@
 Furthermore, using the cumulative number of reported COVID-19 cases, the
-total number of new daily cases is calculated. This variable
-(log-transformed after adding a small constant) is paired with the
+incidence is calculated after dividing by the population of the state
+(in 100,000s). This variable (log-transformed) is paired with the
 corresponding lagged moving average of the mobility indicators. The
-log-transformation is useful to avoid negative values of daily new cases
-when making predictions. Table \ref{tab:descriptive-statistics} shows
-the descriptive statistics of the data set. Analysis is based on
-correlation analysis, multivariate regression, and data visualization.
+log-transformation is useful to avoid negative values of incidence when
+making predictions. Table \ref{tab:descriptive-statistics} shows the
+descriptive statistics of the data set. Analysis is based on correlation
+analysis, multivariate regression, and data visualization.
 
@@
 
-Table \ref{tab:correlation-analysis} shows that the mobility indicators
-are highly correlated with each other. Two variables are selected for
-multivariate analysis: parks- and work-related mobility. Work has a high
-correlation with the outcome variable, and its correlation with parks is
-relatively weak, which increases the information content of the two
-variables. Furthermore, parks- and work-related mobility represent two
-dimensions of out-of-home activities: mandatory and discretionary
-travel.
+Table \ref{tab:correlation-analysis-log-incidence} shows that the
+mobility indicators are highly correlated with each other. Two variables
+are selected for multivariate analysis: parks- and work-related
+mobility. Work has a high correlation with the outcome variable, and its
+correlation with parks is relatively weak, which increases the
+information content of the two variables in multivariate analysis.
+Furthermore, parks- and work-related mobility represent two dimensions
+of out-of-home activities: mandatory and discretionary travel.
 
-\begin{table}[H]
+\begin{table}
 
-\caption{\label{tab:check-correlations}\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators}
+\caption{\label{tab:check-correlations-log-incidence}\label{tab:correlation-analysis-log-incidence}Simple correlation between log(incidence) and the mobility indicators}
 \centering
@@
 \toprule
-  & log\_new\_cases & retail & groceries & parks & transit & work & residential\\
+  & log\_incidence & retail & groceries & parks & transit & work & residential\\
 \midrule
-\rowcolor{gray!6}  log\_new\_cases & 1.00 & -0.68 & -0.48 & -0.27 & -0.68 & -0.69 & 0.71\\
-retail & -0.68 & 1.00 & 0.87 & 0.41 & 0.93 & 0.98 & -0.98\\
-\rowcolor{gray!6}  groceries & -0.48 & 0.87 & 1.00 & 0.44 & 0.87 & 0.87 & -0.87\\
-parks & -0.27 & 0.41 & 0.44 & 1.00 & 0.49 & 0.39 & -0.41\\
-\rowcolor{gray!6}  transit & -0.68 & 0.93 & 0.87 & 0.49 & 1.00 & 0.93 & -0.95\\
+\rowcolor{gray!6}  log\_incidence & 1.00 & -0.86 & -0.73 & -0.20 & -0.82 & -0.91 & 0.89\\
+retail & -0.86 & 1.00 & 0.88 & 0.38 & 0.91 & 0.96 & -0.98\\
+\rowcolor{gray!6}  groceries & -0.73 & 0.88 & 1.00 & 0.41 & 0.88 & 0.87 & -0.88\\
+parks & -0.20 & 0.38 & 0.41 & 1.00 & 0.46 & 0.33 & -0.36\\
+\rowcolor{gray!6}  transit & -0.82 & 0.91 & 0.88 & 0.46 & 1.00 & 0.92 & -0.93\\
 \addlinespace
-work & -0.69 & 0.98 & 0.87 & 0.39 & 0.93 & 1.00 & -0.99\\
-\rowcolor{gray!6}  residential & 0.71 & -0.98 & -0.87 & -0.41 & -0.95 & -0.99 & 1.00\\
+work & -0.91 & 0.96 & 0.87 & 0.33 & 0.92 & 1.00 & -0.98\\
+\rowcolor{gray!6}  residential & 0.89 & -0.98 & -0.88 & -0.36 & -0.93 & -0.98 & 1.00\\
 \bottomrule
@@
 
-A regression model is estimated with the log of new daily cases as the
+A regression model is estimated with the log of incidence as the
 dependent variable. The covariates enter the regression in the form of a
@@
 April 5) is introduced to account for the temporal trend of the
-pandemic. Finally, an indicator variable for the state of New York and
-an interaction with the date are used to distinguish the unusually high
-incidence of the disease there. The results appear in Table
-\ref{tab:model-results}. The model provides a good fit to the data and
-all variables reported are significant at \(p<0.05\) or better.
+pandemic. Finally, an indicator variable for the state of New York is
+used to distinguish the unusually high incidence of the disease there.
+The results of the model are shown in Table
+\ref{tab:model-results-log-incidence}. The model provides a good fit to
+the data and all variables reported are significant at \(p<0.10\) or
+better.
 
-There is an overall temporal trend that indicates a growing number of
-cases over time, at an accelerating rate (see positive sign of
-date\^{}2). Mobility related to parks and to work both tend to increase
-the number of new cases; the negative sign for their interaction is
-indicative of the trade-offs between these two forms of mobility and
-their impact on new cases. The influence of parks-related mobility was
-relatively weaker early in the pandemic (negative sign of the parks x
-date term) but has become more important over time (positive sign of the
-parks x date\^{}2 term). The opposite happens with work-related
-mobility, which started with a greater effect (positive sign of work x
-date term), but whose impact has declined over time (negative sign of
-work x date\^{}2 term) New York has on average more new daily cases than
-the rest of the states, but this has declined over time (see negative
-sign of NY x date).
+There is an overall temporal trend that indicates a growing incidence
+over time, but at a decelerating rate (see negative sign of date\^{}2).
+Mobility related to parks and to work are both associated with higher
+incidence of COVID-19, however, the effect of parks-related mobility
+grows non-linearly (see positive sign of quadratic term), whereas the
+effect of work-related mobility grows at a decreasing rate (see negative
+sign of quadratic term). Furthermore, the negative sign for the
+interaction of these two mobility indicators captures the trade-offs
+between these two forms of mobility and their impact on incidence. The
+influence of parks-related mobility was relatively weak early in the
+pandemic (negative sign of the parks x date term) but has become more
+important over time (positive sign of the parks x date\^{}2 term). The
+opposite happens with work-related mobility, the importance of which has
+declined over time (negative sign of work x date\^{}2 term), but whose
+impact has declined over time (negative sign of work x date\^{}2 term).
+As seen in the table, incidence of COVID-19 in New York is consistently
+higher.
 
@@
 
-\caption{\label{tab:model-results}\label{tab:model-results}Results of estimating regression model. Dependent variable is log(New Cases + 0.0001).}
+\caption{\label{tab:model-results-log-incidence}\label{tab:model-results}Results of estimating regression model. Dependent variable is log(Incidence).}
 \centering
@@
 \midrule
-\rowcolor{gray!6}  date & 0.1085 & <0.001\\
-date\textasciicircum{}2 & 0.0070 & <0.001\\
-\rowcolor{gray!6}  parks\textasciicircum{}2 & 2.1756 & <0.001\\
-parks & 8.9203 & <0.001\\
-\rowcolor{gray!6}  parks x work & -23.6010 & <0.001\\
+\rowcolor{gray!6}  date & 0.1618 & <0.001\\
+date\textasciicircum{}2 & -0.0008 & 0.0093\\
+\rowcolor{gray!6}  parks\textasciicircum{}2 & 0.2518 & 0.0562\\
+parks & 4.1307 & <0.001\\
+\rowcolor{gray!6}  parks x work & -8.1480 & <0.001\\
 \addlinespace
-work & 6.3156 & 0.0035\\
-\rowcolor{gray!6}  work\textasciicircum{}2 & 12.0539 & <0.001\\
-parks x date & -0.2133 & <0.001\\
-\rowcolor{gray!6}  parks x date\textasciicircum{}2 & 0.0079 & <0.001\\
-work x date & 0.1376 & 0.0272\\
+work & 9.4568 & <0.001\\
+\rowcolor{gray!6}  work\textasciicircum{}2 & -3.4421 & <0.001\\
+parks x date & -0.0948 & <0.001\\
+\rowcolor{gray!6}  parks x date\textasciicircum{}2 & 0.0036 & <0.001\\
+work x date\textasciicircum{}2 & -0.0058 & <0.001\\
 \addlinespace
-\rowcolor{gray!6}  work x date\textasciicircum{}2 & -0.0238 & <0.001\\
-NY & 3.5466 & <0.001\\
-\rowcolor{gray!6}  NY x date & -0.0527 & <0.001\\
+\rowcolor{gray!6}  NY & 1.8627 & <0.001\\
 \bottomrule
 \multicolumn{3}{l}{\textit{Note: }}\\
-\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.831}\\
-\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.83}\\
-\multicolumn{3}{l}{Standard Error $\sigma$= 2.081}\\
+\multicolumn{3}{l}{Coefficient of Determination $R^2$= 0.972}\\
+\multicolumn{3}{l}{Adjusted Coefficient of Determination $R^2$= 0.972}\\
+\multicolumn{3}{l}{Standard Error $\sigma$= 0.689}\\
 \end{tabular}
@@
 according to the mobility indicators and date. Figure
-\ref{fig:prediction-plots} shows the prediction surfaces on three
-different dates: March 21, when the first states began implementing
-stay-at-home orders; April 5, fifteen days later; and April 20, fifteen
-more days later, and at a time when some states started to consider
-relaxing stay-at-home orders.
+\ref{fig:prediction-plots} shows the prediction surfaces on four
+different dates at intervals of 15 days: March 21, when the first states
+began implementing stay-at-home orders; then April 5, two weeks into the
+lockdown; this is followed by April 20, at a time when some states
+started to consider relaxing stay-at-home orders; and finally May 5,
+when some states were reopening and/or letting stay-at-home orders
+lapse.
 
-On March 21 there were still only minor departures from baseline
-mobility (recall that these are temporally lagged); the prediction
-surface at this point is relatively flat. This changes by April 5, when
-work-based mobility has declined substantially. Although every state
-registers lower work-based mobility, there are large variations in
-parks-based mobility, with some states seeing increases of up to 60\%
-for this class of mobility.
+On March 21 there were still only minor departures from the baseline
+level of mobility (recall that these are temporally lagged); the
+prediction surface at this point is relatively flat. This changes by
+April 5, when work-based mobility has declined substantially. Although
+every state registers lower work-based mobility, there are large
+variations in parks-based mobility, with some states seeing increases of
+up to 60\% for this class of mobility. By May 5, park-related mobility
+in some states had increased to 200\% of the baseline.
 
-The prediction surface indicates an expectation of a greater number of
-new daily cases as either class of mobility increases, but the effect is
-not linear. On the last date examined, April 20, the trend becomes more
-steep for park-based mobility, even as this indicator continues to
-display large variations from the baseline in both directions. The white
-dashed lines in the plots are the folds of the saddles, and represent,
-for each date, the combination of parks- and work-related mobility
-levels that tended to minimize the emergence of new cases.
+The prediction surfaces are hyperbolic paraboloids on any given date,
+and in general indicate an expectation of higher incidences as either
+class of mobility increases, but with a progressively steeper trend for
+park-based mobility over time. On the last date examined, May 5, the
+trend becomes more steep for park-based mobility, even as this indicator
+continues to display large variations from the baseline in both
+directions. The white dashed lines in the plots are the folds of the
+saddles, and represent, for each date, the combination of parks- and
+work-related mobility levels that tended to minimize the incidence.
 
@@
 The results suggest that over time the benefits of reduced work-related
-mobility can be offset by parks-related mobility. For example, on March
-21, New Jersey and Idaho had similar levels of park-related mobility. By
-April 20, New Jersey had substantially reduced park-related mobility,
-whereas Idaho's was even higher than in March. The estimated and actual
-number of new cases grew in the intervening period; however, New
-Jersey's growth in new cases (from March 21 to April 20) was only 894\%,
-whereas Idaho's was 1030\%.
+mobility can be easily offset by parks-related mobility. For example,
+California has consistently registered lower levels of parks-related
+mobility whereas Idaho has had high levels of this kind of mobility
+throughout the pandemic. The incidence of COVID-19 grew in the
+intervening period; however, between March 21 and May 5 growth in
+incidence in California was 876.42\% whereas Idaho's growth in incidence
+over the same period was 2519.05\%.
 
-These results suggest the potential of GCMR to investigate the spread of
-COVID-19, but also point at some limitations. The baseline level is not
-defined in a metric that is amenable to policy development (e.g.,
-person-km travelled). Without a clearer understanding of the absolute
-levels of these variables, their use can suggest trends, but their
-potential for applied policy analysis appears to be more limited.
+These results suggest the potential of GCMR to investigate the potential
+effects of mobility on the incidence of COVID-19. In particular, growth
+appears to be more strongly driven by parks-related mobility. In terms
+of the use of these mobility indicators, there are some limitations that
+must be acknowledged. The baseline level is not defined in a metric that
+is amenable to policy development (e.g., person-km travelled). Without a
+clearer understanding of the absolute levels of these variables, these
+indicators are useful for inference and perhaps short-term forecasting,
+but their potential for applied policy analysis appears to be more
+limited.
 

### 2020-05-19 README.Rmd
Fixed typo in README.Rmd
v3 Revisions after second round of anonymous reviews.

@@
 
-# Using Google Community Mobility Reports to investigate the growth of COVID-19 in the United States
+# Using Google Community Mobility Reports to investigate the incidence of COVID-19 in the United States
 
@@
 
-(revised versin of paper after anonymous reviews submitted to Transport Findings on May 11, 2020)
+(third version of paper after second round of anonymous reviews; resubmitted to Transport Findings on May 19, 2020)
 
@@
 These reports are based on the company's location-tracking capabilities and measure changes in mobility with respect to a baseline.
-This novel source of data offers an opportunity to investigate potential correlations between mobility and transmission of COVID-19.
-Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and new daily reported cases of COVID-19 by state in the US.
+This novel source of data offers an opportunity to investigate potential correlations between mobility and incidence of COVID-19.
+Using data from the New York Times on COVID-19 cases and GCMR, this paper presents an analysis of mobility levels and incidence of COVID-19 by state in the US.
 The results provide insights about the utility and interpretability of GCMR for COVID-19 research and decision-making.
@@
                   as.character(round(sd(c19$residential), 2)))) %>%
-  kable(#"latex",
-        "html",
+  kable("latex",
         booktabs = TRUE,
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
-  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
   column_spec(2, width = "12em") %>%
@@
   cor(use = "pairwise.complete.obs") %>%
-  kable(#"latex",
-        "html",
+  kable("latex",
         booktabs = TRUE,
@@
         caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
-  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
   mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
-  kable(#"latex",
-        "html",
+  kable("latex",
         booktabs = TRUE,
@@
         col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
-  kable_styling(latex_options = c("striped", "HOLD_position")) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
+  footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
+                       paste("Adjusted Coefficient of Determination $R^2$=", round(summary(mod1)$adj.r.squared, 3)),
+                       paste("Standard Error $\\\\sigma$=", round(summary(mod1)$sigma, 3))),
+           escape = FALSE)
+```
+
+```{r model-results-log-incidence, echo = FALSE}
+data.frame(Variable = c("date",
+                        "date^2",
+                        "parks^2",
+                        "parks",
+                        "parks x work",
+                        "work",
+                        "work^2",
+                        "parks x date",
+                        "parks x date^2",
+                        "work x date^2",
+                        "NY"),
+           Coefficient = as.numeric(coefficients(mod1)),
+           pval = as.numeric(summary(mod1)$coefficients[,4])) %>%
+  mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
+  kable("latex",
+        booktabs = TRUE,
+        digits = 4,
+        caption = "\\label{tab:model-results}Results of estimating regression model. Dependent variable is log(Incidence).",
+        align = c("l", "c", "c"),
+        col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
   footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
@@
 Visualization is the most effective way to understand the trend according to the mobility indicators and date.
-Figure \ref{fig:prediction-plots} shows the prediction surfaces on three different dates: March 21, when the first states began implementing stay-at-home orders; April 5, fifteen days later; and April 20, fifteen more days later, and at a time when some states started to consider relaxing stay-at-home orders.
+Figure \ref{fig:prediction-plots} shows the prediction surfaces on four different dates at intervals of 15 days: March 21, when the first states began implementing stay-at-home orders; then April 5, two weeks into the lockdown; this is followed by April 20, at a time when some states started to consider relaxing stay-at-home orders; and finally May 5, when some states were reopening and/or letting stay-at-home orders lapse.
 
-On March 21 there were still only minor departures from baseline mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
+On March 21 there were still only minor departures from the baseline level of mobility (recall that these are temporally lagged); the prediction surface at this point is relatively flat.
 This changes by April 5, when work-based mobility has declined substantially.
 Although every state registers lower work-based mobility, there are large variations in parks-based mobility, with some states seeing increases of up to 60% for this class of mobility.
+By May 5, park-related mobility in some states had increased to 200% of the baseline.
 
-The prediction surface indicates an expectation of a greater number of new daily cases as either class of mobility increases, but the effect is not linear.
-On the last date examined, April 20, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
-The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the emergence of new cases.
+The prediction surfaces are hyperbolic paraboloids on any given date, and in general indicate an expectation of higher incidences as either class of mobility increases, but with a progressively steeper trend for park-based mobility over time.
+On the last date examined, May 5, the trend becomes more steep for park-based mobility, even as this indicator continues to display large variations from the baseline in both directions.
+The white dashed lines in the plots are the folds of the saddles, and represent, for each date, the combination of parks- and work-related mobility levels that tended to minimize the incidence.
 

### 2020-05-19 README.Rmd
v3 Recreated README.Rmd

@@
                   as.character(round(sd(c19$residential), 2)))) %>%
-  kable("latex",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
         caption = "\\label{tab:descriptive-statistics}Descriptive statistics of the data set") %>%
-  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
   column_spec(2, width = "12em") %>%
@@
         caption = "\\label{tab:correlation-analysis}Simple correlation between log(New Cases) and the mobility indicators")  %>%
-  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position"), font_size = 7) %>%
+  kable_styling(latex_options = c("striped", "scale_down", "HOLD_position")) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
         caption = "\\label{tab:correlation-analysis-log-incidence}Simple correlation between log(incidence) and the mobility indicators")  %>%
-  kable_styling(latex_options = c("striped", "scale_down"), font_size = 7) %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
   footnote(general = "All mobility indicators are lagged 11-day moving averages")
@@
   mutate(pval = ifelse(pval >= 0.001, round(pval, 4), "<0.001")) %>%
-  kable("latex",
+  kable(#"latex",
+        "html",
         booktabs = TRUE,
@@
         col.names = c("Variable", "Coefficient Estimate", "p-value")) %>%
-  kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7) %>%
+  kable_styling(latex_options = c("striped", "HOLD_position")) %>%
   footnote(general = c(paste("Coefficient of Determination $R^2$=", round(summary(mod1)$r.squared, 3)),
